---
title: "Hands-on Exercise 2 - Beyond ggplot2 Fundamentals"
author: "Yap Jit Wei"
date: "April 23, 2026"
date-modified: "last-modified"

format:
  html:
    toc: true
    toc-depth: 4
    number-sections: true
    
execute:
  warning: false 
  freeze: false
---

# Beyond ggplot2 Fundamentals

## Overview

In this chapter, you will be introduced to several ggplot2 extensions for creating more elegant and effective statistical graphics. By the end of this exercise, you will be able to:

-   control the placement of annotation on a graph by using functions provided in ggrepel package,
-   create professional publication quality figure by using functions provided in ggthemes and hrbrthemes packages,
-   plot composite figure by combining ggplot2 graphs by using patchwork package.

## Getting Started

### Installing and loading the required libraries

In this exercise, beside tidyverse, four R packages will be used. They are:

-   ggrepel: an R package provides geoms for ggplot2 to repel overlapping text labels.
-   ggthemes: an R package provides some extra themes, geoms, and scales for ‘ggplot2’.
-   hrbrthemes: an R package provides typography-centric themes and theme components for ggplot2.
-   patchwork: an R package for preparing composite figure created using ggplot2.

Code chunk below will be used to check if these packages have been installed and also will load them onto your working R environment.

```{r}
pacman::p_load(ggrepel, patchwork, 
               ggthemes, hrbrthemes,
               tidyverse) 
```

### Importing Data

For the purpose of this exercise, a data file called Exam_data will be used. It consists of year end examination grades of a cohort of primary 3 students from a local school. It is in csv file format.

The code chunk below imports exam_data.csv into R environment by using [read_csv()](https://readr.tidyverse.org/reference/read_delim.html) function of readr package. [readr](https://readr.tidyverse.org/) is one of the tidyverse package.

```{r}
exam_data <- read_csv("Data/Exam_data.csv")
```
There are a total of seven attributes in the exam_data tibble data frame. Four of them are categorical data type and the other three are in continuous data type.

-   The categorical attributes are: ID, CLASS, GENDER and RACE.
-   The continuous attributes are: MATHS, ENGLISH and SCIENCE.

## Beyond ggplot2 Annotation: ggrepel

One of the challenge in plotting statistical graph is annotation, especially with large number of data points.

::: {.panel-tabset}
### The plot
```{r}
#| echo: false
ggplot(data=exam_data, 
       aes(x= MATHS, 
           y=ENGLISH)) +
  geom_point() +
  geom_smooth(method=lm, 
              size=0.5) +  
  geom_label(aes(label = ID), 
             hjust = .5, 
             vjust = -.5) +
  coord_cartesian(xlim=c(0,100),
                  ylim=c(0,100)) +
  ggtitle("English scores versus Maths scores for Primary 3")
```
### The code

```{r}
#| eval: false
ggplot(data=exam_data, 
       aes(x= MATHS, 
           y=ENGLISH)) +
  geom_point() +
  geom_smooth(method=lm, 
              size=0.5) +  
  geom_label(aes(label = ID), 
             hjust = .5, 
             vjust = -.5) +
  coord_cartesian(xlim=c(0,100),
                  ylim=c(0,100)) +
  ggtitle("English scores versus Maths scores for Primary 3")
```

:::

[ggrepel](https://ggrepel.slowkow.com/) is an extension of ggplot2 package which provides geoms for ggplot2 to repel overlapping text as in our examples on the right.

We simply replace geom_text() by [geom_text_repel()](https://ggrepel.slowkow.com/reference/geom_text_repel.html) and geom_label() by [geom_label_repel()](https://ggrepel.slowkow.com/reference/geom_text_repel.html).

### Working with ggrepel

::: {.panel-tabset}
#### The plot
```{r}
#| echo: false
ggplot(data=exam_data, 
       aes(x= MATHS, 
           y=ENGLISH)) +
  geom_point() +
  geom_smooth(method=lm, 
              size=0.5) +  
  geom_label_repel(aes(label = ID), 
                   fontface = "bold") +
  coord_cartesian(xlim=c(0,100),
                  ylim=c(0,100)) +
  ggtitle("English scores versus Maths scores for Primary 3")
```

#### The code
```{r}
#| eval: false
ggplot(data=exam_data, 
       aes(x= MATHS, 
           y=ENGLISH)) +
  geom_point() +
  geom_smooth(method=lm, 
              size=0.5) +  
  geom_label_repel(aes(label = ID), 
                   fontface = "bold") +
  coord_cartesian(xlim=c(0,100),
                  ylim=c(0,100)) +
  ggtitle("English scores versus Maths scores for Primary 3")
```

:::

#### Extra alternative: Using geom_text_repel()

This alternative uses geom_text_repel() with smaller and lighter labels on the scatterplot. The rest of the chart remains the same as the original example.

This is useful because it reduces visual heaviness and makes the plot look less cluttered.

```{r}
ggplot(data = exam_data,
       aes(x = MATHS,
           y = ENGLISH)) +
  geom_point() +
  geom_smooth(method = lm,
              size = 0.5) +
  geom_text_repel(aes(label = ID),
                  size = 3,
                  color = "grey30") +
  coord_cartesian(xlim = c(0,100),
                  ylim = c(0,100)) +
  ggtitle("English scores versus Maths scores for Primary 3")
```


## Beyond ggplot2 Themes

ggplot2 comes with [eight built-in themes](https://ggplot2.tidyverse.org/reference/ggtheme.html), they are: theme_gray(), theme_bw(), theme_classic(), theme_dark(), theme_light(), theme_linedraw(), theme_minimal(), and theme_void().

::: {.panel-tabset}
#### The plot
```{r}
#| echo: false
ggplot(data=exam_data, 
             aes(x = MATHS)) +
  geom_histogram(bins=20, 
                 boundary = 100,
                 color="grey25", 
                 fill="grey90") +
  theme_gray() +
  ggtitle("Distribution of Maths scores") 
```
#### The code
```{r}
#| eval: false
ggplot(data=exam_data, 
             aes(x = MATHS)) +
  geom_histogram(bins=20, 
                 boundary = 100,
                 color="grey25", 
                 fill="grey90") +
  theme_gray() +
  ggtitle("Distribution of Maths scores") 
```

:::

Refer to this link to learn more about ggplot2 [Themes](https://ggplot2.tidyverse.org/reference/index.html#themes)

### Working with ggtheme package

[ggthemes](https://cran.r-project.org/web/packages/ggthemes/index.html) provides [‘ggplot2’ themes](https://yutannihilation.github.io/allYourFigureAreBelongToUs/ggthemes/) that replicate the look of plots by Edward Tufte, Stephen Few, Fivethirtyeight, [The Economist](https://www.economist.com/topics/graphic-detail), ‘Stata’, ‘Excel’, and [The Wall Street Journal](https://www.pinterest.com/wsjgraphics/wsj-graphics/), among others.

In the example below, The Economist theme is used.

::: {.panel-tabset}
#### The plot
```{r}
#| echo: false
ggplot(data=exam_data, 
             aes(x = MATHS)) +
  geom_histogram(bins=20, 
                 boundary = 100,
                 color="grey25", 
                 fill="grey90") +
  ggtitle("Distribution of Maths scores") +
  theme_economist()
```
#### The code
```{r}
#| eval: false
ggplot(data=exam_data, 
             aes(x = MATHS)) +
  geom_histogram(bins=20, 
                 boundary = 100,
                 color="grey25", 
                 fill="grey90") +
  ggtitle("Distribution of Maths scores") +
  theme_economist()
```
:::

It also provides some extra geoms and scales for ‘ggplot2’. Consult this [vignette](https://yutannihilation.github.io/allYourFigureAreBelongToUs/ggthemes/) to learn more.

#### Extra alternative: Applying wsj()

This alternative applies theme_fivethirtyeight() from the ggthemes package to the scatterplot. It gives the chart a cleaner look with a style similar to Wall Street Journal graphics.

This is useful because it changes the overall visual style of the chart and can cater to different tastes.

```{r}
ggplot(data=exam_data, 
             aes(x = MATHS)) +
  geom_histogram(bins=20, 
                 boundary = 100,
                 color="grey25", 
                 fill="grey90") +
  ggtitle("Distribution of Maths scores") +
  theme_wsj()
```


### Working with hrbthems package

[hrbrthemes](https://cran.r-project.org/web/packages/hrbrthemes/index.html) package provides a base theme that focuses on typographic elements, including where various labels are placed as well as the fonts that are used.

::: {.panel-tabset}
#### The plot
```{r}
#| echo: false
ggplot(data=exam_data, 
             aes(x = MATHS)) +
  geom_histogram(bins=20, 
                 boundary = 100,
                 color="grey25", 
                 fill="grey90") +
  ggtitle("Distribution of Maths scores") +
  theme_ipsum()
```

#### The code
```{r}
#| eval: false
ggplot(data=exam_data, 
             aes(x = MATHS)) +
  geom_histogram(bins=20, 
                 boundary = 100,
                 color="grey25", 
                 fill="grey90") +
  ggtitle("Distribution of Maths scores") +
  theme_ipsum()
```
:::

The second goal centers around productivity for a production workflow. In fact, this “production workflow” is the context for where the elements of hrbrthemes should be used. Consult this [vignette](https://hrbrmstr.github.io/hrbrthemes/articles/why_hrbrthemes.html) to learn more.

::: {.panel-tabset}
#### The plot
```{r}
#| echo: false
ggplot(data=exam_data, 
             aes(x = MATHS)) +
  geom_histogram(bins=20, 
                 boundary = 100,
                 color="grey25", 
                 fill="grey90") +
  ggtitle("Distribution of Maths scores") +
  theme_ipsum(axis_title_size = 18,
              base_size = 15,
              grid = "Y")
```

#### The code
```{r}
#| eval: false
ggplot(data=exam_data, 
             aes(x = MATHS)) +
  geom_histogram(bins=20, 
                 boundary = 100,
                 color="grey25", 
                 fill="grey90") +
  ggtitle("Distribution of Maths scores") +
  theme_ipsum(axis_title_size = 18,
              base_size = 15,
              grid = "Y")
```
::: 

::: {.callout-tip}
## What can we learn from the code chunk above?

- `axis_title_size` argument is used to increase the font size of the axis title to 18,
- `base_size` argument is used to increase the default axis label to 15, and
- `grid` argument is used to remove the x-axis grid lines.
:::

## Beyond Single Graph

It is not unusual that multiple graphs are required to tell a compelling visual story. There are several ggplot2 extensions provide functions to compose figure with multiple graphs. In this section, you will learn how to create composite plot by combining multiple graphs. First, let us create three statistical graphics by using the code chunk below.

::: {.panel-tabset}
#### The plot
```{r}
#| echo: false
p1 <- ggplot(data=exam_data, 
             aes(x = MATHS)) +
  geom_histogram(bins=20, 
                 boundary = 100,
                 color="grey25", 
                 fill="grey90") + 
  coord_cartesian(xlim=c(0,100)) +
  ggtitle("Distribution of Maths scores")

p1
```

#### The code
```{r}
#| eval: false
p1 <- ggplot(data=exam_data, 
             aes(x = MATHS)) +
  geom_histogram(bins=20, 
                 boundary = 100,
                 color="grey25", 
                 fill="grey90") + 
  coord_cartesian(xlim=c(0,100)) +
  ggtitle("Distribution of Maths scores")
```

:::

Next

::: {.panel-tabset}
#### The plot
```{r}
#| echo: false
p2 <- ggplot(data=exam_data, 
             aes(x = ENGLISH)) +
  geom_histogram(bins=20, 
                 boundary = 100,
                 color="grey25", 
                 fill="grey90") +
  coord_cartesian(xlim=c(0,100)) +
  ggtitle("Distribution of English scores")

p2
```

#### The code
```{r}
#| eval: false
p2 <- ggplot(data=exam_data, 
             aes(x = ENGLISH)) +
  geom_histogram(bins=20, 
                 boundary = 100,
                 color="grey25", 
                 fill="grey90") +
  coord_cartesian(xlim=c(0,100)) +
  ggtitle("Distribution of English scores")
```
:::

Lastly, we will draw a scatterplot for English score versus Maths score by as shown below

::: {.panel-tabset}
#### The plot
```{r}
#| echo: false
p3 <- ggplot(data=exam_data, 
             aes(x= MATHS, 
                 y=ENGLISH)) +
  geom_point() +
  geom_smooth(method=lm, 
              size=0.5) +  
  coord_cartesian(xlim=c(0,100),
                  ylim=c(0,100)) +
  ggtitle("English scores versus Maths scores for Primary 3")

p3
```

#### The code
```{r}
#| eval: false
p3 <- ggplot(data=exam_data, 
             aes(x= MATHS, 
                 y=ENGLISH)) +
  geom_point() +
  geom_smooth(method=lm, 
              size=0.5) +  
  coord_cartesian(xlim=c(0,100),
                  ylim=c(0,100)) +
  ggtitle("English scores versus Maths scores for Primary 3")
```
:::

### Creating Composite Graphics: pathwork methods

There are several ggplot2 extension’s functions support the needs to prepare composite figure by combining several graphs such as [grid.arrange()](https://cran.r-project.org/web/packages/gridExtra/vignettes/arrangeGrob.html) of gridExtra package and [plot_grid()](https://wilkelab.org/cowplot/reference/plot_grid.html) of [cowplot](https://wilkelab.org/cowplot/index.html) package. In this section, I am going to shared with you an ggplot2 extension called [patchwork](https://patchwork.data-imaginist.com/) which is specially designed for combining separate ggplot2 graphs into a single figure.

Patchwork package has a very simple syntax where we can create layouts super easily. Here’s the general syntax that combines:

-   Two-Column Layout using the Plus Sign +.
-   Parenthesis () to create a subplot group.
-   Two-Row Layout using the Division Sign /

### Combining two ggplot2 graphs

Figure in the tabset below shows a composite of two histograms created using patchwork. Note how simple the syntax used to create the plot!

::: {.panel-tabset}
#### The plot
```{r}
#| echo: false
p1 + p2

```

#### The code
```{r}
#| eval: false
p1 + p2

```
:::

### Combining three ggplot2 graphs

We can plot more complex composite by using appropriate operators. For example, the composite figure below is plotted by using:

-   “/” operator to stack two ggplot2 graphs,
-   “|” operator to place the plots beside each other,
-   “()” operator the define the sequence of the plotting.

::: {.panel-tabset}
#### The plot
```{r}
#| echo: false
(p1 / p2) | p3


```

#### The code
```{r}
#| eval: false
(p1 / p2) | p3


```
:::

To learn more about, refer to [Plot Assembly](https://patchwork.data-imaginist.com/articles/guides/assembly.html).

#### Extra alternative: Changing patchwork layout arrangement

This alternative changes the layout of the composite figure using patchwork by arranging the plots in a different structure.

This is useful because it allows more flexible control over how multiple plots are presented together.

```{r}
(p1 | p2) / p3

```


### Creating a composite figure with tag

In order to identify subplots in text, patchwork also provides auto-tagging capabilities as shown in the figure below.

::: {.panel-tabset}
#### The plot
```{r}
#| echo: false
((p1 / p2) | p3) + 
  plot_annotation(tag_levels = 'I')


```

#### The code
```{r}
#| eval: false
((p1 / p2) | p3) + 
  plot_annotation(tag_levels = 'I')


```
:::





### Creating figure with insert

Beside providing functions to place plots next to each other based on the provided layout. With [inset_element()](https://patchwork.data-imaginist.com/reference/inset_element.html) of patchwork, we can place one or several plots or graphic elements freely on top or below another plot.

::: {.panel-tabset}
#### The plot
```{r}
#| echo: false
p3 + inset_element(p2, 
                   left = 0.02, 
                   bottom = 0.7, 
                   right = 0.5, 
                   top = 1)


```

#### The code
```{r}
#| eval: false
p3 + inset_element(p2, 
                   left = 0.02, 
                   bottom = 0.7, 
                   right = 0.5, 
                   top = 1)

```
:::

### Creating a composite figure by using patchwork and ggtheme

Figure below is created by combining patchwork and theme_economist() of ggthemes package discussed earlier.

::: {.panel-tabset}
#### The plot
```{r}
#| echo: false
patchwork <- (p1 / p2) | p3
patchwork & theme_economist()


```

#### The code
```{r}
#| eval: false
patchwork <- (p1 / p2) | p3
patchwork & theme_economist()

```
:::

## Reference

-   [Patchwork R package goes nerd viral](https://www.littlemissdata.com/blog/patchwork)
-   [ggrepel](https://ggrepel.slowkow.com/)
-   [ggthemes](https://ggplot2.tidyverse.org/reference/ggtheme.html)
-   [hrbrthemes](https://cinc.rud.is/web/packages/hrbrthemes/)
-   [ggplot tips: Arranging plots](https://albert-rapp.de/post/2021-10-28-extend-plot-variety/)
-   [ggplot2 Theme Elements Demonstration](https://henrywang.nl/ggplot2-theme-elements-demonstration/)
-   [ggplot2 Theme Elements Reference Sheet](https://isabella-b.com/blog/ggplot2-theme-elements-reference/)


