---
title: "Hands-on Exercise 5a - Creating Ternary Plot with R"
author: "Yap Jit Wei"
date: "May 4, 2026"
date-modified: "last-modified"

format:
  html:
    toc: true
    toc-depth: 4
    number-sections: true
    
execute:
  warning: false 
  freeze: auto
---

# Creating Ternary Plot with R

## Overview

Ternary plots are a way of displaying the distribution and variability of three-part compositional data. (For example, the proportion of aged, economy active and young population or sand, silt, and clay in soil.) It’s display is a triangle with sides scaled from 0 to 1. Each side represents one of the three components. A point is plotted so that a line drawn perpendicular from the point to each leg of the triangle intersect at the component values of the point.

In this hands-on, you will learn how to build ternary plot programmatically using R for visualising and analysing population structure of Singapore.

The hands-on exercise consists of four steps:

-   Install and launch **tidyverse** and **ggtern** packages.

-   Derive three new measures using *mutate()* function of **dplyr** package.

-   Build a static ternary plot using *ggtern()* function of **ggtern** package.

-   Build an interactive ternary plot using *plot-ly()* function of **Plotly R** package.

## Installing and launching R packages

For this exercise, two main R packages will be used in this hands-on exercise, they are:

-   [**ggtern**](http://www.ggtern.com/), a ggplot extension specially designed to plot ternary diagrams. The package will be used to plot static ternary plots.

-   [**Plotly R**](https://plot.ly/r/), an R package for creating interactive web-based graphs via plotly’s JavaScript graphing library, plotly.js . The **plotly R** libary contains the *ggplotly* function, which will convert **ggplot2** figures into a Plotly object.

We will also need to ensure that selected **tidyverse** family packages namely: **readr**, **dplyr** and **tidyr** are also installed and loaded.

The code chunks below will accomplish the task.

```{r}
pacman::p_load(plotly, ggtern, tidyverse)
```

## Data Preparation

### The data

For the purpose of this hands-on exercise, the [Singapore Residents by Planning AreaSubzone, Age Group, Sex and Type of Dwelling, June 2000-2018](https://www.singstat.gov.sg/find-data/search-by-theme/population/geographic-distribution/latest-data) data will be used. The data set has been downloaded and included in the data sub-folder of the hands-on exercise folder. It is called *respopagsex2000to2018_tidy.csv* and is in csv file format.

### Importing Data

To important *respopagsex2000to2018_tidy.csv* into R, ***read_csv()*** function of **readr** package will be used.

```{r}
#Reading the data into R environment
pop_data <- read_csv("Data/respopagsex2000to2018_tidy.csv") 
```

### Preparing Data

Next, use the ***mutate()*** function of **dplyr** package to derive three new measures, namely: young, active, and old.

```{r}
#Deriving the young, economy active and old measures
agpop_mutated <- pop_data %>%
  mutate(`Year` = as.character(Year))%>%
  spread(AG, Population) %>%
  mutate(YOUNG = rowSums(.[4:8]))%>%
  mutate(ACTIVE = rowSums(.[9:16]))  %>%
  mutate(OLD = rowSums(.[17:21])) %>%
  mutate(TOTAL = rowSums(.[22:24])) %>%
  filter(Year == 2018)%>%
  filter(TOTAL > 0)
```

## Plotting Ternary Diagram with R

### Plotting a static ternary diagram

Use ***ggtern()*** function of **ggtern** package to create a simple ternary plot.

```{r}
#Building the static ternary plot
ggtern(data=agpop_mutated,aes(x=YOUNG,y=ACTIVE, z=OLD)) +
  geom_point()
```

```{r}
#Building the static ternary plot 
ggtern(data=agpop_mutated, aes(x=YOUNG,y=ACTIVE, z=OLD)) + geom_point() + labs(title="Population structure, 2018") + theme_rgbw()
```

### Plotting an interactive ternary diagram

The code below create an interactive ternary plot using ***plot_ly()*** function of **Plotly R**.

```{r}
# reusable function for creating annotation object
label <- function(txt) {
  list(
    text = txt, 
    x = 0.1, y = 1,
    ax = 0, ay = 0,
    xref = "paper", yref = "paper", 
    align = "center",
    font = list(family = "serif", size = 15, color = "white"),
    bgcolor = "#b3b3b3", bordercolor = "black", borderwidth = 2
  )
}

# reusable function for axis formatting
axis <- function(txt) {
  list(
    title = txt, tickformat = ".0%", tickfont = list(size = 10)
  )
}

ternaryAxes <- list(
  aaxis = axis("Young"), 
  baxis = axis("Active"), 
  caxis = axis("Old")
)

# Initiating a plotly visualization 
plot_ly(
  agpop_mutated, 
  a = ~YOUNG, 
  b = ~ACTIVE, 
  c = ~OLD, 
  color = I("black"), 
  type = "scatterternary"
) %>%
  layout(
    annotations = label("Ternary Markers"), 
    ternary = ternaryAxes
  )
```

## Extra Alternative

### Animated Population Structure (2000–2018)
```{r}
agpop_full <- pop_data %>%
  mutate(Year = as.character(Year)) %>%
  spread(AG, Population) %>%
  mutate(YOUNG  = rowSums(.[4:8]),
         ACTIVE = rowSums(.[9:16]),
         OLD    = rowSums(.[17:21]),
         TOTAL  = YOUNG + ACTIVE + OLD) %>%
  filter(TOTAL > 0)

plot_ly(
  agpop_full,
  a = ~YOUNG,
  b = ~ACTIVE,
  c = ~OLD,
  frame = ~Year,
  type = "scatterternary",
  mode = "markers",
  marker = list(size = 6, opacity = 0.6, color = "steelblue")
) %>%
  layout(
    title = list(
      text = "Singapore Population Structure, 2000–2018",
      x = 0,
      y = 0.95,
      font = list(size = 14)
    ),
    ternary = ternaryAxes
  ) %>%
  animation_opts(frame = 800, transition = 400, easing = "linear") %>%
  animation_slider(currentvalue = list(prefix = "Year: "))
```
The exercise creates a static ternary diagram for 2018 only. This alternative animates the same plot across all 19 years from 2000 to 2018 using frame = ~Year, allowing readers to watch how Singapore's population structure shifts over time. The steady movement of points toward the OLD corner as the years progress makes the ageing trend immediately visible in a way that a single year snapshot cannot show.
