---
title: "Take Home Exercise 1"
subtitle: "Diagnosing Loss Drivers in a Spanish Motor Portfolio"
author: "Yap Jit Wei"
date: "May 20, 2026"
date-modified: "last-modified"

format:
  html:
    toc: true
    toc-depth: 4
    number-sections: true
    
execute:
  warning: false 
  freeze: auto
---



# Overview

## Setting the scene

Property and casualty insurance portfolios routinely underperform without the insurer being able to diagnose why. Loss-driving pockets hide inside coarse segments. Mispriced coverage tiers persist because no one has visualised the loss-ratio gap. Premium inadequacy in narrow sub-books goes undetected until a bad year exposes it.

::: callout-note
"Insurers often carry underperforming portfolios where the root causes of poor profitability and high volatility are unknown. Leveraging visual analytics can potentially help investigate trends in underlying loss drivers, while data enrichment using external data can help refine segmentation and underwriting strategy."

Swiss Re Institute (2019), *sigma 4/2019: Advanced analytics: unlocking new frontiers in P&C insurance*
:::

Motor insurance is a particularly fertile setting for visual analytics. Premiums are exposure-weighted. Claims arrive frequently enough to support statistical inference at the segment level. A loss-ratio improvement of two to five percentage points, achievable through better segmentation alone, can move a motor book between profit and loss.

## Analytical objectives

This report applies visual analytics to a Spanish motor vehicle insurance portfolio with two objectives:

1.  **Univariate analysis**: characterise the distribution of the portfolio's numerical and categorical variables, surfacing concentration, skewness, and structural features that inform downstream segmentation.

2.  **Bivariate analysis with statistical validation**: investigate the factors influencing portfolio profitability (proxied by loss ratio) and volatility (proxied by the coefficient of variation of loss ratio across policies within a segment), with statistical validation of the observed patterns.

Beyond this scope, the analysis applies two visual analytics methods that sit outside Lessons 1 to 5, selected to leverage features of the dataset that standard univariate and bivariate methods cannot reach.

## Dataset

The data is a real-world Spanish motor insurance portfolio published through Mendeley Data, comprising 354,140 policy-year records across 185,678 unique policyholders between 2022 and 2024. The dataset contains 47 variables organised into four logical groups:

-   **Policy and insured characteristics**: `policy_type`, `bonus_score`, `business_type`, `payment_frequency`, and the observation year.
-   **Driver and vehicle characteristics**: `driver_age`, `vehicle_age`, `fuel_type`, `vehicle_value`, `power_to_weight_ratio`, `circulation_area`, and others.
-   **Premiums**: `total_premium` and seven coverage-specific premiums (liability, property damage, theft, fire, glass, legal protection, occupants).
-   **Exposure, claim counts, and incurred costs**: total and per coverage.

The granularity of premium and incurred figures permits coverage-specific loss-ratio decomposition in addition to portfolio-level analysis. This level of detail is uncommon in publicly available motor insurance datasets, and the report exploits it where it adds analytical leverage.

::: callout-note
**Reproducibility**: The raw data file (`Dataset of motor insurance portfolio.csv`, approximately 90MB) exceeds GitHub's recommended file size limit and is therefore excluded from this repository via `.gitignore`. To reproduce the analysis, download the dataset from Mendeley Data and place it in the `Data/` folder at the project root. The analytical sandbox (`motor_clean.rds`) is generated on first render by the code in Section 2 and does not need to be committed.
:::

## Report structure

Section 2 documents the analytical sandbox preparation: import, cleaning, type encoding, and derivation of the profitability and volatility variables. Section 3 presents the univariate distributional analysis. Section 4 examines bivariate relationships with statistical validation of the observed effects. Section 5 introduces two visual analytics methods that extend beyond the techniques covered in class. Section 6 summarises the key findings and identifies directions for future research.

# Analytical sandbox preparation

## Loading R packages

The analytical pipeline relies on the tidyverse for data wrangling, several ggplot2 extensions for visualisation, and supporting packages for table presentation. Each package is selected for a specific role. The rationale is summarised below.

| Package       | Purpose                                                |
|---------------|--------------------------------------------------------|
| `tidyverse`   | Core wrangling and visualisation grammar               |
| `readxl`      | Reading the data dictionary `.xlsx`                    |
| `janitor`     | Variable name cleaning and duplicate detection         |
| `skimr`       | Compact summary of numerical and categorical variables |
| `naniar`      | Visualising missingness patterns                       |
| `ggstatsplot` | Statistical test annotations on plots                  |
| `ggdist`      | Uncertainty and distribution geoms                     |
| `ggridges`    | Ridgeline density plots                                |
| `patchwork`   | Composite plot layouts                                 |
| `scales`      | Axis formatting (percent, comma, currency)             |
| `knitr`       | Table rendering                                        |

```{r}
pacman::p_load(tidyverse, readxl, janitor, skimr, naniar,
               ggstatsplot, ggdist, ggridges,
               patchwork, scales, knitr)
```

```{r}
# House style for all charts 
theme_house <- function(base_size = 11) {
  theme_minimal(base_size = base_size) +
    theme(
      plot.title = element_text(face = "bold", size = base_size + 2),
      plot.subtitle = element_text(colour = "grey40",
                                   size = base_size - 1,
                                   margin = margin(b = 10)),
      plot.caption = element_text(colour = "grey50",
                                  size = base_size - 3,
                                  hjust = 0),
      panel.grid.major.x = element_blank(),
      panel.grid.minor = element_blank(),
      axis.text = element_text(size = base_size - 2),
      axis.title = element_text(size = base_size - 1.5),
      axis.title.x = element_text(margin = margin(t = 8)),
      axis.title.y = element_text(margin = margin(r = 8)),
      plot.margin = margin(t = 10, r = 15, b = 15, l = 10),
      legend.position = "top",
      legend.title = element_text(size = base_size - 2, face = "bold"),
      legend.text = element_text(size = base_size - 2)
    )
}

# House colour palette
house_highlight <- "#C44E52"      # red — for the one important data point
house_muted     <- "grey75"       # for everything else
house_benchmark <- "grey40"       # for reference lines and benchmarks
house_sequential <- "RdYlBu"      # for diverging palettes (loss ratio gradients)
```



## Importing the raw data

The portfolio file is a semicolon-delimited CSV using European decimal conventions. The companion data dictionary is a small Excel workbook documenting variable meanings and units, used as a reference during encoding decisions.

```{r}
raw <- read_delim("Data/Dataset of motor insurance portfolio.csv",
                  delim = ";", show_col_types = FALSE) |>
  clean_names()

dict <- read_excel("Data/Descriptive of variables.xlsx") |>
  clean_names()
```

`clean_names()` from **janitor** standardises column names to snake_case, removing the need for backtick-escaping downstream.

## Initial inspection

```{r}
glimpse(raw)
```

The portfolio contains 354,140 policy-year records across 47 variables. Each record is one policy-year, meaning a single policyholder may appear in multiple rows if their policy spans 2022, 2023, and 2024. This panel structure is preserved for the bivariate analysis in Section 4 and exploited for the bonus analysis in Section 5.

### Duplicates

```{r}
raw |> 
  get_dupes() |> 
  nrow()
```

No exact duplicates were detected.

### Missingness

```{r}
#| fig-cap: "Missing value patterns across the portfolio."
gg_miss_var(raw, show_pct = TRUE)
```

Four variables contain missing values. `fuel_type` (1,287 missing, 0.36%) and `vehicle_value` (513 missing, 0.14%) account for nearly all of it. The remaining columns have two or fewer NAs each. Because missingness is concentrated in non-target descriptive variables and represents under half a percent of records, listwise deletion is acceptable without introducing meaningful selection bias.

```{r}
raw <- raw |> drop_na(fuel_type, vehicle_value)
```

The two patterns of missingness are non-overlapping. No row has both `fuel_type` and `vehicle_value` missing, so listwise deletion removes exactly 1,800 records (0.51% of the portfolio) without any compounding effect.

## Variable encoding

Categorical variables are imported as character vectors. Converting them to factors with explicit, meaningful level ordering improves both downstream plot readability and statistical test behaviour.

Two ordering choices warrant explanation. The `policy_type` levels are ordered from least to most comprehensive coverage (TP, TPG, CC, COMP_E, COMP_N), enabling monotonic interpretation of loss-ratio gradients across coverage tiers. The `bonus_score` levels are ordered from best to worst claim history (G, N, B), following the bonus-malus convention used by motor insurers.

```{r}
clean <- raw |>
  mutate(
    year             = as.integer(year),
    policy_type      = factor(policy_type,
                              levels = c("TP","TPG","CC","COMP_E","COMP_N")),
    policy_status    = factor(policy_status, levels = c("A","C"),
                              labels = c("Active","Cancelled")),
    business_type    = factor(business_type, levels = c("NB","P"),
                              labels = c("New business","Persistent")),
    payment_frequency = factor(payment_frequency, levels = c("A","S","Q"),
                               labels = c("Annual","Semi-annual","Quarterly")),
    bonus_score      = factor(bonus_score, levels = c("G","N","B"),
                              labels = c("Good","Neutral","Bad")),
    fuel_type        = factor(fuel_type, levels = c("D","G"),
                              labels = c("Diesel","Gasoline")),
    municipality_type = factor(municipality_type, levels = c("I","C","IS"),
                               labels = c("Interior","Coastal","Islands")),
    circulation_area = factor(circulation_area, levels = c("U","R"),
                              labels = c("Urban","Rural"))
  )
```

## Derived analytical variables

The brief identifies portfolio profitability and volatility as the outcomes of interest, but neither exists in the raw data. Both must be derived from the relationship between premiums collected, losses incurred, and policy exposure. Three policy-level metrics form the foundation:

-   **Loss ratio**: incurred losses divided by earned premium. The industry-standard measure of pure underwriting profitability before expense loading. A loss ratio below approximately 70% indicates a contribution to operating profit on a motor book.
-   **Claim frequency**: claim count divided by exposure (in policy-years). Exposure-normalising is essential because most policies in this dataset are on-risk for only part of the year (mean exposure 0.71).
-   **Claim severity**: incurred losses divided by claim count, defined only where claims occurred.

```{r}
clean <- clean |>
  mutate(
    loss_ratio      = if_else(total_premium > 0,
                              total_incurred / total_premium, NA_real_),
    claim_frequency = if_else(total_exposure > 0,
                              total_claims / total_exposure, NA_real_),
    claim_severity  = if_else(total_claims > 0,
                              total_incurred / total_claims, NA_real_)
  )
```

Three additional banded variables are constructed to support bivariate segmentation. Bands use breaks informed by the underlying distributions (verified in Section 3) and motor underwriting practice.

```{r}
clean <- clean |> 
  mutate(
    driver_age_band = cut(driver_age,
                          breaks = c(0, 25, 35, 50, 65, Inf),
                          labels = c("<25","25-34","35-49","50-64","65+"),
                          right = FALSE),
    vehicle_age_band = cut(vehicle_age,
                           breaks = c(-1, 2, 5, 10, 15, Inf),
                           labels = c("0-2","3-5","6-10","11-15","16+"),
                           right = TRUE),
    vehicle_value_band = cut(vehicle_value,
                             breaks = quantile(vehicle_value,
                                               probs = c(0, .25, .5, .75, 1),
                                               na.rm = TRUE),
                             labels = c("Q1","Q2","Q3","Q4"),
                             include.lowest = TRUE)
  )
```

`driver_age_band` follows actuarial convention: the under-25 band is consistently the highest-risk group in motor insurance, and 65+ is the standard senior threshold. `vehicle_age_band` uses breaks aligned with depreciation milestones. `vehicle_value_band` uses quartiles rather than fixed cuts because vehicle values span several orders of magnitude.

## Volatility metric

Volatility is operationalised at the segment level as the coefficient of variation (CV) of loss ratio across the policies within a segment. Unlike standard deviation, CV is scale-free, making it directly comparable across segments with very different loss-ratio means. A high CV indicates that a segment's profitability is unstable. Even if its mean loss ratio is acceptable, the distribution of individual-policy outcomes is wide, raising the risk of unexpected losses in any given year.

This summary is computed on demand within the bivariate analysis (Section 4) rather than added to the policy-level data frame, as it is inherently a group-level statistic.

## Saving the analytical sandbox

The cleaned, encoded, and enriched dataset is written to disk as a single `.rds` file. The slides deliverable consumes the same file, ensuring perfect consistency between report and presentation.

```{r}
write_rds(clean, "Data/motor_clean.rds")
```

```{r}
clean |> 
  summarise(
    records = n(),
    policyholders = n_distinct(insured_id),
    years_covered = paste(range(year), collapse = "–"),
    total_premium = sum(total_premium),
    total_incurred = sum(total_incurred),
    portfolio_loss_ratio = total_incurred / total_premium
  ) |>
  kable(caption = "Analytical sandbox summary",
        format.args = list(big.mark = ","))
```

## Wrangling flowchart

The figure below summarises the analytical-sandbox preparation pipeline.

```{mermaid}
flowchart LR
  A[Raw CSV<br/>354,140 × 47] --> B[clean_names<br/>snake_case]
  D[Data dictionary<br/>.xlsx] -.reference.-> B
  B --> C[Drop NA<br/>fuel_type, vehicle_value]
  C --> E[Factor encoding<br/>8 categoricals]
  E --> F[Derive KPIs<br/>loss_ratio, frequency, severity]
  F --> G[Band continuous<br/>driver/vehicle age, value]
  G --> H[(motor_clean.rds<br/>352,340 × 53)]
```

# Univariate visual analysis

## Methodology

Univariate analysis serves two purposes in this exercise. First, it documents the shape of each variable, including central tendency, spread, skewness, and modality, which constrains the techniques available for bivariate analysis. Second, it surfaces concentrations and outliers that often turn out to be the loss drivers the brief asks the analyst to identify.

Two visual conventions are used:

-   **Continuous variables**: a histogram paired with a horizontal boxplot via the `patchwork` package. The histogram shows distribution shape. The boxplot makes outliers and the five-number summary explicit. The composite is a more honest representation than either alone.
-   **Categorical variables**: an ordered bar chart with frequency labels. Ordering by frequency (rather than alphabetical or factor-default) reveals concentration immediately, with the most common category at the top.

```{r}
clean <- read_rds("Data/motor_clean.rds")
```

## Continuous variables

### Driver age

```{r}
#| fig-height: 4

h_age <- ggplot(clean, aes(x = driver_age)) +
  geom_histogram(bins = 40, fill = "#4C72B0", colour = "white") +
  scale_y_continuous(labels = comma, expand = expansion(mult = c(0, 0.05))) +
  labs(x = NULL, y = "Number of policies") +
  theme_house(base_size = 11) +
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank())

b_age <- ggplot(clean, aes(x = driver_age)) +
  geom_boxplot(fill = "#4C72B0", colour = "grey30",
               width = 0.5, outlier.alpha = 0.1) +
  scale_y_continuous(breaks = NULL) +
  labs(x = "Driver age (years)", y = NULL) +
  theme_house(base_size = 11) +
  theme(axis.text.y = element_blank())

(h_age / b_age + plot_layout(heights = c(4, 1))) +
  plot_annotation(
    title = "Driver age ranges from 22 to 82, centred in mid-40s",
    subtitle = "Distribution of driver age across 352,340 policy-years",
    caption = "Source: Spanish motor insurance portfolio (Mendeley Data).",
    theme = theme(
      plot.title = element_text(face = "bold", size = 13),
      plot.subtitle = element_text(colour = "grey40", size = 10,
                                   margin = margin(b = 10)),
      plot.caption = element_text(colour = "grey50", size = 8, hjust = 0)
    )
  )
```

The driver-age distribution spans 22 to 82 years, centred in the mid-40s, with a modest right tail of older drivers. The under-25 segment is small in volume but structurally important: it is the conventional high-risk pool in motor underwriting and becomes a focal point of the bivariate analysis. The 65+ tail is similarly relevant, as senior drivers carry distinct accident-risk profiles.

### Vehicle age

```{r}
#| fig-height: 4

h_va <- ggplot(clean, aes(x = vehicle_age)) +
  geom_histogram(bins = 30, fill = "#4C72B0", colour = "white") +
  scale_y_continuous(labels = comma, expand = expansion(mult = c(0, 0.05))) +
  labs(x = NULL, y = "Number of policies") +
  theme_house(base_size = 11) +
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank())

b_va <- ggplot(clean, aes(x = vehicle_age)) +
  geom_boxplot(fill = "#4C72B0", colour = "grey30",
               width = 0.5, outlier.alpha = 0.1) +
  scale_y_continuous(breaks = NULL) +
  labs(x = "Vehicle age (years)", y = NULL) +
  theme_house(base_size = 11) +
  theme(axis.text.y = element_blank())

(h_va / b_va + plot_layout(heights = c(4, 1))) +
  plot_annotation(
    title = "Most insured vehicles are between 18 and 36 years old",
    subtitle = "Distribution of vehicle age across 352,340 policy-years",
    caption = "Source: Spanish motor insurance portfolio (Mendeley Data).",
    theme = theme(
      plot.title = element_text(face = "bold", size = 13),
      plot.subtitle = element_text(colour = "grey40", size = 10,
                                   margin = margin(b = 10)),
      plot.caption = element_text(colour = "grey50", size = 8, hjust = 0)
    )
  )
```

The vehicle-age distribution spans 0 to about 60 years, with most vehicles concentrated between 18 and 36 years old (the IQR) and a median around 25 years. This is an unusually aged portfolio: the bulk of insured vehicles are over 15 years old, with a long tail of vehicles older than 40. The shape suggests two underwriting considerations. Theft exposure tends to concentrate in newer, higher-value vehicles, of which this portfolio has few. Breakdown and mechanical-failure claims rise with vehicle age, which this portfolio has in abundance. These tensions surface in the bivariate analysis of coverage tier and in the driver-age × vehicle-age risk surface.

### Vehicle value

```{r}
#| fig-height: 4

h_vv <- ggplot(clean, aes(x = vehicle_value)) +
  geom_histogram(bins = 50, fill = "#4C72B0", colour = "white") +
  scale_x_log10(labels = label_currency(prefix = "€", scale = 1e-3,
                                         suffix = "k")) +
  scale_y_continuous(labels = comma, expand = expansion(mult = c(0, 0.05))) +
  labs(x = NULL, y = "Number of policies") +
  theme_house(base_size = 11) +
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank())

b_vv <- ggplot(clean, aes(x = vehicle_value)) +
  geom_boxplot(fill = "#4C72B0", colour = "grey30",
               width = 0.5, outlier.alpha = 0.1) +
  scale_x_log10(labels = label_currency(prefix = "€", scale = 1e-3,
                                         suffix = "k")) +
  scale_y_continuous(breaks = NULL) +
  labs(x = "Vehicle value (€, log scale)", y = NULL) +
  theme_house(base_size = 11) +
  theme(axis.text.y = element_blank())

(h_vv / b_vv + plot_layout(heights = c(4, 1))) +
  plot_annotation(
    title = "Vehicle value clusters around €20-30k with a heavy upper tail",
    subtitle = "Distribution of vehicle value across 352,340 policy-years (log scale)",
    caption = "Source: Spanish motor insurance portfolio (Mendeley Data).\nLog scale used to reveal the central distribution; the small fraction of high-value vehicles extends beyond €100k.",
    theme = theme(
      plot.title = element_text(face = "bold", size = 13),
      plot.subtitle = element_text(colour = "grey40", size = 10,
                                   margin = margin(b = 10)),
      plot.caption = element_text(colour = "grey50", size = 8, hjust = 0)
    )
  )
```

Vehicle value is right-skewed, with the bulk of policies covering vehicles worth between €20,000 and €30,000 and a small share covering higher-value vehicles above €100,000. The log-scale histogram makes the central concentration visible; on a linear scale, the rare high-value outliers would compress the typical distribution into the left edge. This concentration in mid-value vehicles is consistent with the older fleet composition described above.

### Power-to-weight ratio

```{r}
h_pw <- ggplot(clean, aes(x = power_to_weight_ratio)) +
  geom_histogram(bins = 50, fill = "#4C72B0", colour = "white") +
  scale_x_log10() +
  scale_y_continuous(labels = comma, expand = expansion(mult = c(0, 0.05))) +
  labs(x = NULL, y = "Number of policies") +
  theme_house(base_size = 11) +
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank())

b_pw <- ggplot(clean, aes(x = power_to_weight_ratio)) +
  geom_boxplot(fill = "#4C72B0", colour = "grey30",
               width = 0.5, outlier.alpha = 0.1) +
  scale_x_log10() +
  scale_y_continuous(breaks = NULL) +
  labs(x = "Power-to-weight ratio (log scale)", y = NULL) +
  theme_house(base_size = 11) +
  theme(axis.text.y = element_blank())

(h_pw / b_pw + plot_layout(heights = c(4, 1))) +
  plot_annotation(
    title = "Most vehicles have a power-to-weight ratio between 10 and 15",
    subtitle = "Distribution of power-to-weight ratio across 352,340 policy-years (log scale)",
    caption = "Source: Spanish motor insurance portfolio (Mendeley Data).",
    theme = theme(
      plot.title = element_text(face = "bold", size = 13),
      plot.subtitle = element_text(colour = "grey40", size = 10,
                                   margin = margin(b = 10)),
      plot.caption = element_text(colour = "grey50", size = 8, hjust = 0)
    )
  )
```

Power-to-weight ratio is right-skewed on the raw scale and approximately log-normal once transformed. Most vehicles fall in the 10-15 range, with a long upper tail of high-performance vehicles and a smaller tail at the low end. The upper tail carries the underwriting risk concern: power-to-weight correlates with claim frequency and severity in motor portfolios, and a small number of high-performance policies can disproportionately affect segment-level loss ratios.

## Categorical variables

### Policy type

```{r}
#| fig-height: 3.5

ggplot(clean, aes(x = fct_infreq(policy_type))) +
  geom_bar(fill = "#4C72B0", colour = "white") +
  geom_text(stat = "count",
            aes(label = comma(after_stat(count))),
            vjust = -0.4, size = 3.2, colour = "grey30") +
  scale_y_continuous(labels = comma, expand = expansion(mult = c(0, 0.15))) +
  labs(
    title = "Most policies cluster in the mid-comprehensive tier (CC)",
    subtitle = "Distribution of Policy Type across 352,340 policy-years",
    x = NULL,
    y = "Number of policies",
    caption = "Source: Spanish motor insurance portfolio (Mendeley Data)."
  ) +
  theme_house(base_size = 11)
```

The portfolio is dominated by the two combined tiers: CC (204,318 policies, around 58%) and COMP_E (90,320 policies, around 26%). The third-party tiers (TP and TPG) together account for around 12% of policies, and COMP_N (comprehensive without excess) is a small but consequential segment at around 5%. As Section 4 shows, COMP_N carries by far the highest loss intensity relative to premium charged, while CC and COMP_E sit closer to the portfolio benchmark.

### Bonus-Malus Category

```{r}
#| fig-height: 3.5

ggplot(clean, aes(x = fct_infreq(bonus_score))) +
  geom_bar(fill = "#4C72B0", colour = "white") +
  geom_text(stat = "count",
            aes(label = comma(after_stat(count))),
            vjust = -0.4, size = 3.2, colour = "grey30") +
  scale_y_continuous(labels = comma, expand = expansion(mult = c(0, 0.15))) +
  labs(
    title = "97% of policy-years are in the Good bonus-malus category",
    subtitle = "Distribution of bonus-malus category across 352,340 policy-years",
    x = NULL,
    y = "Number of policies",
    caption = "Source: Spanish motor insurance portfolio (Mendeley Data)."
  ) +
  theme_house(base_size = 11)
```

The bonus-malus distribution is extreme: 96.5% of policy-years sit in the Good category, with Neutral (2.2%) and Bad (1.3%) together accounting for the remaining 3.5%. The two smaller categories are sufficient in absolute volume for statistical testing (several thousand records each) but small in relative terms. This imbalance becomes important in Section 4.4, where the analysis shows that the small Neutral pool carries a disproportionate share of the portfolio's loss-ratio deterioration.

### Other categoricals

```{r}
#| fig-height: 7

cat_vars <- c("policy_status", "business_type", "payment_frequency",
              "fuel_type", "municipality_type", "circulation_area")

cat_summary <- clean |>
  select(all_of(cat_vars)) |>
  pivot_longer(everything(), names_to = "variable", values_to = "value") |>
  count(variable, value) |>
  group_by(variable) |>
  mutate(pct = n / sum(n)) |>
  ungroup() |>
  mutate(variable = factor(variable,
                            levels = cat_vars,
                            labels = c("Policy status",
                                       "Business type",
                                       "Payment frequency",
                                       "Fuel type",
                                       "Municipality type",
                                       "Circulation area")))

ggplot(cat_summary, aes(x = fct_reorder(value, n), y = n)) +
  geom_col(fill = "#4C72B0") +
  geom_text(aes(label = percent(pct, accuracy = 0.1)),
            hjust = -0.1, size = 2.8, colour = "grey30") +
  scale_y_continuous(labels = comma, expand = expansion(mult = c(0, 0.2))) +
  coord_flip() +
  facet_wrap(~ variable, scales = "free", ncol = 2) +
  labs(
    title = "Most categorical variables are heavily concentrated in their dominant category",
    subtitle = "Distribution of six categorical variables across 352,340 policy-years",
    x = NULL,
    y = "Number of policies",
    caption = "Source: Spanish motor insurance portfolio (Mendeley Data).\nPercentages are within-variable; bars within each facet are sorted by frequency."
  ) +
  theme_house(base_size = 11) +
  theme(
    strip.text = element_text(face = "bold", colour = "grey20", size = 10),
    strip.background = element_rect(fill = "grey95", colour = NA),
    panel.grid.major.y = element_blank(),
    panel.spacing = unit(1.2, "lines")
  )
```

Several concentrations are worth noting. The portfolio is predominantly new business (69%) rather than persistent renewals (31%), pointing to rapid book growth across the period. New business typically carries higher loss ratios than renewals because insurers price renewals more precisely from claims history, so a fast-growing book can deteriorate even without underlying segment changes. Annual payment dominates over instalment options. Diesel vehicles outnumber gasoline by roughly two to one. Urban and rural circulation are close to balanced (54% vs 46%), though within-category loss-ratio profiles differ markedly as Section 4.5 shows. Interior locations dominate the geographic distribution (65%), with coastal and islands together accounting for 35%; despite their smaller volume, the islands segment carries the widest within-category loss-ratio spread.

## Distribution of derived KPIs

The three derived KPIs are themselves variables worth profiling. At the policy level, they are highly skewed: most policies have no claims, so most loss ratios and frequencies are zero. Restricting the view to policies with at least one claim reveals the within-distribution shape that matters for segment-level analysis.

```{r}
share_above_be <- clean |>
  filter(total_claims > 0, is.finite(loss_ratio)) |>
  summarise(share = mean(loss_ratio > 1)) |>
  pull(share)

clean |>
  filter(total_claims > 0, is.finite(loss_ratio),
         loss_ratio >= 0.01, loss_ratio <= 100) |>
  ggplot(aes(x = loss_ratio, fill = loss_ratio > 1)) +
  geom_histogram(bins = 50, boundary = 1, colour = "white") +
  scale_x_continuous(
    trans  = "log10",
    breaks = c(0.01, 0.1, 1, 10, 100),
    labels = c("1%", "10%", "100%", "1,000%", "10,000%")
  ) +
  scale_y_continuous(labels = comma, expand = expansion(mult = c(0, 0.12))) +
  scale_fill_manual(
    values = c("FALSE" = house_muted, "TRUE" = house_highlight),
    guide  = "none"
  ) +
  geom_vline(xintercept = 1, linetype = "dashed", colour = house_benchmark) +
  annotate(
    "text",
    x = 1.15, y = Inf,
    label  = "Break-even (100%)",
    hjust  = 0, vjust = 1.2,
    colour = house_benchmark, size = 3.2
  ) +
  labs(
    title    = "A substantial share of claim-bearing policies exceed break-even",
    subtitle = paste0(
      scales::percent(share_above_be, accuracy = 0.1),
      " of claim-bearing policy-years have losses exceeding their premium"
    ),
    x       = "Loss ratio (log scale)",
    y       = "Policy-years",
    caption = "Source: Spanish motor insurance portfolio (Mendeley Data)."
  ) +
  theme_house()
```

Among policies with at least one claim, the loss-ratio distribution spans nearly four orders of magnitude. The dashed grey line marks the break-even point where losses equal premium. The red bars to the right represent policy-years that returned more in claims than they collected in premium. Segment-level mean loss ratios mask this within-policy variance, which is what the volatility (coefficient of variation) analysis in Section 4 surfaces.


## Summary of univariate observations

The univariate view raises three structural features the bivariate analysis must contend with:

1.  **Strong distributional skew** in vehicle value and the derived KPIs. Linear-scale plots will mislead. Log scales and percentile bands are appropriate.
2.  **Heavy class imbalance** in `bonus_score` (concentrated in Good) and `policy_type` (concentrated in CC and COMP_E). Statistical tests must account for unequal group sizes.
3.  **The zero-claim majority**. Roughly 85% of policy-years have no claims, making policy-level loss ratio a noisy individual measure but a clean segment-level one when aggregated by exposure-weighted sums.

## Pairwise relationships among continuous variables

Before turning to bivariate analysis of the categorical risk drivers, the continuous variables are examined jointly through a correlation matrix. The correlogram surfaces three things: which variable pairs move together strongly enough to constrain interpretation downstream, which appear independent and can be treated as separate signals, and whether any pair carries multicollinearity that would distort multivariate inference.

```{r}
if (!requireNamespace("ggcorrplot", quietly = TRUE)) {
  install.packages("ggcorrplot")
}
library(ggcorrplot)
```

```{r}
#| fig-width: 8
#| fig-height: 6.5

cont_vars <- clean |>
  select(driver_age, vehicle_age, age_driving_licence,
         vehicle_value, seats, power_to_weight_ratio,
         total_premium, total_exposure, loss_ratio) |>
  filter(if_all(everything(), is.finite))

corr_matrix <- cor(cont_vars, use = "pairwise.complete.obs",
                   method = "spearman")

p_matrix <- cor_pmat(cont_vars, method = "spearman")

ggcorrplot(corr_matrix,
           type = "lower",
           lab = TRUE,
           lab_size = 3,
           p.mat = p_matrix,
           sig.level = 0.05,
           insig = "blank",
           colors = c("#4C72B0", "white", house_highlight),
           outline.color = "white",
           tl.cex = 9) +
  labs(
    title = "Driver age and licence tenure are the strongest pair",
    subtitle = "Spearman correlations among continuous variables; non-significant cells (p ≥ 0.05) blanked",
    x = NULL, y = NULL,
    fill = "Spearman ρ",
    caption = "Source: Spanish motor insurance portfolio (Mendeley Data).\nSpearman rank correlation used because several variables are heavily skewed."
  ) +
  theme_house(base_size = 10) +
  theme(panel.grid = element_blank(),
        legend.position = "right",
        axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1),
        plot.margin = margin(t = 10, r = 15, b = 30, l = 30))
```

Three patterns are visible. The strongest correlation in the matrix is between driver_age and age_driving_licence (ρ = 0.86), which is structural and expected: older drivers have held licences longer. A second strong association sits between total_premium and total_exposure (ρ = 0.74), also structural, since premium earned scales with the on-risk period of a policy. The most analytically interesting cell is the moderate negative correlation between vehicle_value and power_to_weight_ratio (ρ = -0.5), indicating that higher-value vehicles in this portfolio tend to be heavier sedans rather than performance vehicles, the opposite of the pattern in younger, sportier markets. Beyond these three, all remaining pairs are weak. loss_ratio in particular shows only weak correlations with the underlying predictors, confirming that no single continuous variable explains policy-level loss outcomes on its own. This is the empirical motivation for the segment-level bivariate analysis that follows: where individual continuous variables fail to discriminate, categorical and joint structure may succeed.


# Bivariate visual analysis with statistical validation

## Methodology

The bivariate analysis investigates how the portfolio's policy, vehicle, driver, and geographic characteristics relate to its profitability (loss ratio) and volatility (coefficient of variation of loss ratio across policies within a segment).

Three analytical principles guide the work:

1.  **Segment-level loss ratio, not policy-level**. Individual policies have highly skewed loss ratios, since most are zero and a few are very large. Aggregating to the segment level via `sum(total_incurred) / sum(total_premium)` produces the exposure-weighted metric an underwriter would price against.
2.  **Statistical validation alongside every visual claim**. Where the relationship is categorical, `ggstatsplot::ggbetweenstats` or `ggbarstats` provides inline test annotations (ANOVA, Kruskal-Wallis, chi-square). Where the comparison is across years on a ratio of sums, bootstrap confidence intervals are used because the sampling distribution of a ratio is not well-described by closed-form parametric tests.
3.  **Volatility reported alongside profitability**. A segment with high mean loss ratio is unprofitable. A segment with high CV of loss ratio is unpredictable. Underwriting interventions for these two cases differ and should not be conflated.


## Temporal trajectory

The first question to ask of any insurance portfolio is whether the recent loss experience is stable, improving, or worsening. The aggregate view by year shows clear movement.

```{r}
yr <- clean |>
  group_by(year) |>
  summarise(
    premium  = sum(total_premium),
    incurred = sum(total_incurred),
    exposure = sum(total_exposure),
    claims   = sum(total_claims),
    n_pol    = n(),
    .groups  = "drop"
  ) |>
  mutate(
    loss_ratio = incurred / premium,
    frequency  = claims / exposure,
    severity   = incurred / claims
  )

yr |>
  select(year, n_pol, premium, incurred, loss_ratio, frequency, severity) |>
  kable(
    caption = "Portfolio KPIs by year",
    digits = c(0, 0, 0, 0, 4, 4, 0),
    format.args = list(big.mark = ",")
  )
```

Three movements show up at once. Loss ratio rose from approximately 66% in 2022 to 75% in 2024, a 9-percentage-point swing in two years. Frequency edged up modestly. Severity increased more meaningfully, indicating that average claim cost rather than claim count is the principal driver. To establish whether this trajectory is real or a small-sample artefact, the bootstrap distribution of loss ratio is computed for each year.


```{r}
#| fig-height: 4.5
set.seed(608)
B <- 500

boot_one_year <- function(df, B) {
  n <- nrow(df)
  prem <- df$total_premium
  inc  <- df$total_incurred
  replicate(B, {
    idx <- sample.int(n, n, replace = TRUE)
    sum(inc[idx]) / sum(prem[idx])
  })
}

boot_df <- clean |>
  filter(total_premium > 0) |>
  group_by(year) |>
  group_modify(~ tibble(loss_ratio = boot_one_year(.x, B))) |>
  ungroup()


boot_summary <- boot_df |>
  group_by(year) |>
  summarise(
    lr_median = median(loss_ratio),
    lr_lower = quantile(loss_ratio, 0.025),
    lr_upper = quantile(loss_ratio, 0.975),
    .groups = "drop"
  )

ggplot(boot_df, aes(x = loss_ratio, y = factor(year))) +
  stat_halfeye(aes(fill = factor(year)),
               .width = c(.95), point_interval = "mean_qi",
               alpha = 0.8, slab_colour = "white", slab_linewidth = 0.5) +
  geom_vline(xintercept = 0.70, linetype = "dashed", colour = house_benchmark) +
  annotate("text", x = 0.70, y = 3.55, label = "70% benchmark",
           hjust = 1.05, size = 3, colour = "grey40", fontface = "italic") +
  annotate("curve",
           x = 0.78, xend = 0.745,
           y = 3.2, yend = 3.1,
           arrow = arrow(length = unit(2, "mm"), type = "closed"),
           curvature = 0.3, colour = house_highlight, linewidth = 0.4) +
  annotate("text",
           x = 0.79, y = 3.25,
           label = "2024 distribution sits\nentirely above earlier years",
           hjust = 0, size = 3.2, colour = house_highlight, lineheight = 0.95) +
  scale_x_continuous(labels = label_percent(accuracy = 1),
                     limits = c(0.6, 0.85)) +
  scale_fill_manual(values = c("2022" = "grey75",
                                "2023" = "grey55",
                                "2024" = house_highlight),
                    guide = "none") +
  labs(
    title = "2024 loss ratio is well outside the prior-year range",
    subtitle = "Bootstrap distribution of portfolio loss ratio by year (500 resamples)",
    x = "Bootstrap loss ratio",
    y = NULL,
    caption = "Source: Spanish motor insurance portfolio (Mendeley Data)."
  ) +
  theme_house(base_size = 11)
```

The 95% intervals for 2022 and 2023 overlap substantially, but the 2024 distribution sits entirely above both prior years. The intervals are well-separated and statistically distinct. The deterioration is a real shift, not sampling noise.

```{r}
#| fig-height: 4

yr_per_pol <- yr |>
  mutate(
    premium_per_pol = premium / n_pol,
    incurred_per_pol = incurred / n_pol
  ) |>
  select(year, premium_per_pol, incurred_per_pol) |>
  pivot_longer(c(premium_per_pol, incurred_per_pol),
               names_to = "metric") |>
  mutate(metric = factor(metric,
                          levels = c("premium_per_pol", "incurred_per_pol"),
                          labels = c("Premium per policy (€)",
                                     "Incurred losses per policy (€)")))

ggplot(yr_per_pol, aes(x = year, y = value, colour = metric, group = metric)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 3.5) +
  geom_text(aes(label = paste0("€", round(value))),
            vjust = -1.4, size = 3.2, fontface = "bold", show.legend = FALSE) +
  scale_x_continuous(breaks = c(2022, 2023, 2024)) +
  scale_y_continuous(expand = expansion(mult = c(0.1, 0.25))) +
  scale_colour_manual(values = c("Premium per policy (€)" = house_benchmark,
                                  "Incurred losses per policy (€)" = house_highlight)) +
  labs(
    title = "Incurred losses rose faster than premium across the period",
    subtitle = "Per-policy economics, 2022–2024",
    x = "Year",
    y = "Euros per policy",
    colour = NULL,
    caption = "Source: Spanish motor insurance portfolio (Mendeley Data).\nThe deterioration arises from segment-level mix shift, documented from Section 4.3."
  ) +
  theme_house(base_size = 11) +
  theme(legend.position = "top")
```

The aggregate decomposition shows a clear gap: premium per policy rose from €278 in 2022 to €303 in 2024 (a 9 percent increase), while incurred losses per policy rose from €183 to €226 (a 24 percent increase). Average frequency and average severity remained nearly flat across the period (frequency 0.302 to 0.311, severity €969 to €971). The portfolio-level deterioration therefore did not arise from a uniform rise in claim incidence or claim size. It arose from a mix shift in the portfolio. The remainder of Section 4 identifies the specific segments where premium and exposure diverged, beginning with policy type in Section 4.3.

## Coverage tier: the COMP_N premium-exposure mismatch

The portfolio is segmented into five coverage tiers by `policy_type`, ranging from basic third-party (TP) to comprehensive without excess (COMP_N). Loss-ratio variation across these tiers reveals whether premium charged is commensurate with the broader coverage offered.

```{r}
pt <- clean |>
  group_by(policy_type) |>
  summarise(
    n_pol      = n(),
    premium    = sum(total_premium),
    incurred   = sum(total_incurred),
    exposure   = sum(total_exposure),
    claims     = sum(total_claims),
    loss_ratio = incurred / premium,
    frequency  = claims / exposure,
    severity   = if_else(claims > 0, incurred / claims, NA_real_),
    .groups    = "drop"
  )

pt |>
  kable(
    caption = "Portfolio KPIs by policy type",
    digits = c(0, 0, 0, 0, 0, 0, 4, 4, 0),
    format.args = list(big.mark = ",")
  )
```

The numbers reveal a monotonic gradient. Basic third-party (TP) policies run a loss ratio of around 54%, well inside profitable territory. Loss ratio rises through TPG and CC, reaches the mid-70s for COMP_E, and arrives at COMP_N at approximately 94%. The COMP_N segment is within touching distance of pure-loss breakeven before any expense loading is applied.

```{r}
#| fig-height: 5

pt_plot_data <- pt |>
  mutate(
    highlight = if_else(policy_type == "COMP_N", "COMP_N", "Other tiers"),
    label_text = percent(loss_ratio, accuracy = 0.1)
  )

ggplot(pt_plot_data, aes(x = policy_type, y = loss_ratio, fill = highlight)) +
  geom_col(width = 0.7) +
  geom_hline(yintercept = 0.70, linetype = "dashed", colour = "grey40") +
  geom_text(aes(label = label_text,
                colour = highlight),
            vjust = -0.6, size = 3.6, fontface = "bold") +
  geom_text(aes(label = paste0("n = ", comma(n_pol)), y = 0.02),
            colour = "white", size = 2.8, vjust = 0) +
  annotate("text",
           x = 0.6, y = 0.72,
           label = "70% benchmark",
           hjust = 0, size = 3, colour = "grey40", fontface = "italic") +
  annotate("curve",
           x = 4.4, xend = 4.95,
           y = 0.85, yend = 0.93,
           arrow = arrow(length = unit(2, "mm"), type = "closed"),
           curvature = -0.3, colour = "#C44E52", linewidth = 0.4) +
  annotate("text",
           x = 4.2, y = 0.84,
           label = "COMP_N runs 24 points\nabove the benchmark",
           hjust = 1, size = 3.2, colour = "#C44E52", lineheight = 0.95) +
  scale_y_continuous(labels = label_percent(accuracy = 1),
                     limits = c(0, 1.05),
                     expand = expansion(mult = c(0, 0.05))) +
  scale_fill_manual(values = c("COMP_N" = "#C44E52",
                                "Other tiers" = "grey75"),
                    guide = "none") +
  scale_colour_manual(values = c("COMP_N" = "#C44E52",
                                  "Other tiers" = "grey40"),
                      guide = "none") +
  labs(
    title = "Comprehensive-without-excess runs at 94% loss ratio",
    subtitle = "Loss ratio across five coverage tiers, 2022–2024 portfolio",
    x = "Policy type (ordered from least to most comprehensive)",
    y = "Loss ratio",
    caption = "Source: Spanish motor insurance portfolio (Mendeley Data)."
  ) +
  theme_minimal(base_size = 11) +
  theme(
    plot.title = element_text(face = "bold", size = 13),
    plot.subtitle = element_text(colour = "grey40", size = 10, margin = margin(b = 10)),
    plot.caption = element_text(colour = "grey50", size = 8, hjust = 0),
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank(),
    axis.text = element_text(size = 9),
    axis.title = element_text(size = 9.5)
  )
```

To validate that the policy-type differences reflect underlying segment behaviour rather than artefacts of skewed within-segment distributions, a non-parametric test is applied to the policy-level loss ratio (filtered to policies with claims, where the metric is informative).

Before the formal test, a ridgeline view of the policy-level loss-ratio distributions makes the within-segment shape visible across all five tiers at once.

```{r}
#| fig-height: 4.5

clean |>
  filter(total_claims > 0, is.finite(loss_ratio),
         loss_ratio >= 0.01, loss_ratio <= 100) |>
  ggplot(aes(x = loss_ratio,
             y = fct_rev(policy_type),
             fill = policy_type)) +
  geom_density_ridges(scale = 1.4, alpha = 0.85,
                      colour = "white", linewidth = 0.4,
                      quantile_lines = TRUE, quantiles = 2) +
  geom_vline(xintercept = 1,
             linetype = "dashed", colour = house_benchmark) +
  annotate("text", x = 1.05, y = 5.6,
           label = "Break-even (100%)",
           hjust = 0, size = 3, colour = house_benchmark,
           fontface = "italic") +
  scale_x_continuous(trans = "log10",
                     breaks = c(0.01, 0.1, 1, 10, 100),
                     labels = c("1%", "10%", "100%", "1,000%", "10,000%")) +
  scale_fill_manual(values = c("TP"     = "grey75",
                                "TPG"    = "grey75",
                                "CC"     = "grey75",
                                "COMP_E" = "grey75",
                                "COMP_N" = house_highlight),
                    guide = "none") +
  labs(
    title = "COMP_N's loss-ratio distribution sits visibly to the right of every other tier",
    subtitle = "Ridgeline densities of policy-level loss ratio across five coverage tiers (claims-positive policies, log scale)",
    x = "Loss ratio (log scale)",
    y = NULL,
    caption = "Source: Spanish motor insurance portfolio (Mendeley Data).\nWhite vertical lines mark the within-tier median."
  ) +
  theme_house(base_size = 11) +
  theme(panel.grid.major.y = element_blank(),
        plot.margin = margin(t = 25, r = 15, b = 15, l = 10))
```

```{r}
clean |>
  filter(total_claims > 0,
         is.finite(loss_ratio),
         loss_ratio >= 0.05, loss_ratio <= 50) |>
  ggbetweenstats(
    x = policy_type,
    y = loss_ratio,
    type = "np",
    pairwise.display = "significant",
    p.adjust.method = "fdr",
    plot.type = "boxviolin",
    centrality.plotting = TRUE,
    bf.message = FALSE,
    ggtheme = theme_minimal()
  ) +
  scale_colour_manual(values = c("TP" = "grey60",
                                  "TPG" = "grey60",
                                  "CC" = "grey60",
                                  "COMP_E" = "grey60",
                                  "COMP_N" = house_highlight),
                      guide = "none") +
  scale_fill_manual(values = c("TP" = "grey60",
                                "TPG" = "grey60",
                                "CC" = "grey60",
                                "COMP_E" = "grey60",
                                "COMP_N" = house_highlight),
                    guide = "none") +
  scale_y_continuous(
    trans  = "log10",
    breaks = c(0.1, 0.5, 1, 5, 20),
    labels = c("10%", "50%", "100%", "500%", "2,000%")
  ) +
  coord_cartesian(ylim = c(0.05, 50)) +
  labs(x = "Policy type",
       y = "Policy-level loss ratio (log scale, claims-positive only)")
```

The Kruskal-Wallis test on policy-level loss ratio rejects the null of equal distributions across policy types at any conventional significance level. Pairwise comparisons (FDR-adjusted) confirm that COMP_N is significantly different from every other tier. The finding represents a systematic pricing inadequacy rather than statistical noise.

The diagnosis is that comprehensive-without-excess is sold at premium levels that do not reflect its actual loss exposure. Swiss Re's framework describes this pattern as a premium and exposure mismatch.


## The bonus-malus paradox

`bonus_score` is the insurer's primary mechanism for adjusting premium to historical claim behaviour. Three categories are used: Good (no recent claims), Neutral (intermediate), and Bad (recent claims history). The expectation is straightforward. Claim frequency should rise from Good to Bad, and loss ratio should follow, unless the pricing applied to Bad policies has already absorbed their elevated frequency.

```{r}
bs <- clean |>
  group_by(bonus_score) |>
  summarise(
    n_pol      = n(),
    premium    = sum(total_premium),
    incurred   = sum(total_incurred),
    exposure   = sum(total_exposure),
    claims     = sum(total_claims),
    loss_ratio = incurred / premium,
    frequency  = claims / exposure,
    severity   = if_else(claims > 0, incurred / claims, NA_real_),
    avg_prem   = premium / exposure,
    .groups    = "drop"
  )

bs |>
  kable(
    caption = "Portfolio KPIs by bonus-malus category",
    digits = c(0, 0, 0, 0, 0, 0, 4, 4, 0, 0),
    format.args = list(big.mark = ",")
  )
```

The frequency column behaves as expected: 0.30 for Good, 0.48 for Neutral, 0.62 for Bad, a clean monotonic rise. Premium per exposure also rises monotonically, as the bonus-malus surcharges intend. The loss ratio tells a different story.

```{r}
#| fig-height: 7

bs_se <- clean |>
  group_by(bonus_score) |>
  summarise(
    n_pol      = n(),
    incurred   = sum(total_incurred),
    premium    = sum(total_premium),
    exposure   = sum(total_exposure),
    claims     = sum(total_claims),
    loss_ratio = incurred / premium,
    lr_se      = sqrt((incurred / premium) * (1 - incurred / premium) / premium),
    lr_lower   = pmax(0, loss_ratio - 1.96 * lr_se),
    lr_upper   = loss_ratio + 1.96 * lr_se,
    frequency  = claims / exposure,
    freq_se    = sqrt(claims) / exposure,
    freq_lower = pmax(0, frequency - 1.96 * freq_se),
    freq_upper = frequency + 1.96 * freq_se,
    avg_prem   = premium / exposure,
    .groups    = "drop"
  ) |>
  mutate(highlight = if_else(bonus_score == "Neutral", "Neutral", "Other"))

panel_style <- list(
  scale_colour_manual(values = c("Neutral" = house_highlight,
                                  "Other" = house_benchmark),
                      guide = "none"),
  theme_house(base_size = 10),
  theme(plot.subtitle = element_text(face = "bold", colour = "grey20",
                                     size = 10, margin = margin(b = 5)))
)

p_lr <- ggplot(bs_se, aes(x = bonus_score, y = loss_ratio, colour = highlight)) +
  geom_errorbar(aes(ymin = lr_lower, ymax = lr_upper), width = 0.15, linewidth = 0.6) +
  geom_point(size = 4) +
  geom_text(aes(label = percent(loss_ratio, accuracy = 0.1)),
            vjust = -1.6, size = 3.3, fontface = "bold") +
  scale_y_continuous(labels = label_percent(accuracy = 1),
                     expand = expansion(mult = c(0.15, 0.45))) +
  labs(x = NULL, y = "Loss ratio", subtitle = "Loss ratio (with 95% CI)") +
  panel_style

p_freq <- ggplot(bs_se, aes(x = bonus_score, y = frequency, colour = highlight)) +
  geom_errorbar(aes(ymin = freq_lower, ymax = freq_upper), width = 0.15, linewidth = 0.6) +
  geom_point(size = 4) +
  geom_text(aes(label = round(frequency, 3)),
            vjust = -1.6, size = 3.3, fontface = "bold") +
  scale_y_continuous(expand = expansion(mult = c(0.15, 0.45))) +
  labs(x = NULL, y = "Frequency", subtitle = "Claim frequency (claims per policy-year)") +
  panel_style

p_prem <- ggplot(bs_se, aes(x = bonus_score, y = avg_prem, colour = highlight)) +
  geom_point(size = 4) +
  geom_text(aes(label = paste0("€", round(avg_prem))),
            vjust = -1.6, size = 3.3, fontface = "bold") +
  scale_y_continuous(expand = expansion(mult = c(0.15, 0.45))) +
  labs(x = "Bonus-malus category", y = "Premium / exposure (€)",
       subtitle = "Average premium per unit of exposure") +
  panel_style

(p_lr / p_freq / p_prem) +
  plot_annotation(
    title = "Neutral pays less but loses more: the bonus-malus paradox",
    subtitle = "Loss ratio, frequency, and premium per exposure by bonus-malus category, 2022–2024",
    caption = "Source: Spanish motor insurance portfolio (Mendeley Data). 95% confidence intervals from binomial (loss ratio) and Poisson (frequency) approximations.",
    theme = theme(
      plot.title = element_text(face = "bold", size = 13),
      plot.subtitle = element_text(colour = "grey40", size = 10, margin = margin(b = 12)),
      plot.caption = element_text(colour = "grey50", size = 8, hjust = 0)
    )
  )
```

The three-panel composite makes the paradox visible. Frequency rises monotonically across Good, Neutral, and Bad (0.30, 0.48, 0.62 claims per policy-year), and average premium per exposure does the same (€408, €668, €899). Loss ratio does not follow that pattern. Neutral runs at 81.9 percent, materially above the 79.6 percent of Bad, despite collecting 26 percent less premium per exposure. The 95% confidence intervals on loss ratio and frequency confirm that the gap is not noise. The bonus-malus mechanism is absorbing the elevated frequency in the Bad tier but not in Neutral.

```{r}
clean |>
  filter(total_claims > 0,
         is.finite(loss_ratio),
         loss_ratio >= 0.05, loss_ratio <= 50) |>
  ggbetweenstats(
    x = bonus_score,
    y = loss_ratio,
    type = "np",
    pairwise.display = "significant",
    p.adjust.method = "fdr",
    plot.type = "boxviolin",
    bf.message = FALSE,
    ggtheme = theme_minimal()
  ) +
  scale_colour_manual(values = c("Good" = "grey60",
                                  "Neutral" = house_highlight,
                                  "Bad" = "grey60"),
                      guide = "none") +
  scale_fill_manual(values = c("Good" = "grey60",
                                "Neutral" = house_highlight,
                                "Bad" = "grey60"),
                    guide = "none") +
  scale_y_continuous(
    trans  = "log10",
    breaks = c(0.1, 0.5, 1, 5, 20),
    labels = c("10%", "50%", "100%", "500%", "2,000%")
  ) +
  coord_cartesian(ylim = c(0.05, 50)) +
  labs(x = "Bonus-malus category",
       y = "Policy-level loss ratio (log scale, claims-positive only)")
```

The chart shows policy-level loss ratio among claims-positive policies, which inverts the directional pattern of the segment-level aggregate above. Good-tier policies have the lowest premium, so a single claim produces a high policy-level ratio; the segment-level view is dominated instead by the much larger share of zero-claim policies. The statistical test confirms that the Neutral and Bad distributions differ, and that Good is materially separated from both. The Neutral segment is the pricing gap. Two non-exclusive explanations are plausible:

1.  **New entrants land in Neutral by default**, before sufficient claim history has accumulated to move them up or down, and the default premium does not load adequately for the embedded risk in this group.
2.  **Bonus-malus transitions are downward-rigid**: a policyholder who claims may stay in Neutral rather than move to Bad, again undercharging for actual loss potential.

Section 5 (panel transition analysis) examines the second hypothesis directly by tracking individual policyholders across years.

## Geographic interaction: where losses concentrate

The portfolio is described by two geographic variables. `municipality_type` distinguishes interior, coastal, and island regions. `circulation_area` classifies the policy's primary driving environment as urban or rural. Examined in isolation, neither variable produces a dramatic loss-ratio gradient. Their interaction does.

```{r}
geo <- clean |>
  group_by(municipality_type, circulation_area) |>
  summarise(
    n_pol      = n(),
    premium    = sum(total_premium),
    incurred   = sum(total_incurred),
    exposure   = sum(total_exposure),
    claims     = sum(total_claims),
    loss_ratio = incurred / premium,
    frequency  = claims / exposure,
    .groups    = "drop"
  )

geo |>
  kable(
    caption = "Portfolio KPIs by municipality type and circulation area",
    digits = c(0, 0, 0, 0, 0, 0, 0, 4, 4),
    format.args = list(big.mark = ",")
  )
```

```{r}
#| fig-height: 4.5

geo_segment <- clean |>
  group_by(municipality_type, circulation_area) |>
  summarise(
    n_pol = n(),
    premium = sum(total_premium),
    incurred = sum(total_incurred),
    loss_ratio = incurred / premium,
    .groups = "drop"
  ) |>
  mutate(
    is_outlier = municipality_type == "Islands" & circulation_area == "Urban"
  )

ggplot(geo_segment,
       aes(x = circulation_area, y = municipality_type, fill = loss_ratio)) +
  geom_tile(colour = "white", linewidth = 1.5) +
  geom_text(aes(label = paste0(percent(loss_ratio, accuracy = 0.1),
                                "\nn = ", comma(n_pol)),
                fontface = if_else(is_outlier, "bold", "plain")),
            size = 3.4, lineheight = 0.95,
            colour = "grey20") +
    scale_fill_distiller(palette = house_sequential, direction = -1,
                       labels = label_percent(accuracy = 1),
                       name = "Loss ratio") +
  scale_x_discrete(expand = c(0, 0)) +
  scale_y_discrete(expand = c(0, 0)) +
  coord_cartesian(clip = "off", ylim = c(0.5, 3.5)) +
  labs(
    title = "Urban island policies run 11 points above rural island policies",
    subtitle = "Loss ratio by circulation area and municipality type",
    x = "Circulation area",
    y = "Municipality type",
    caption = "Source: Spanish motor insurance portfolio (Mendeley Data).\nCell labels show loss ratio and number of policies."
  ) +
  theme_house(base_size = 11) +
  theme(
    panel.grid = element_blank(),
    legend.position = "right",
    plot.margin = margin(t = 35, r = 15, b = 15, l = 50)
  )
```

The heatmap shows the interaction at a glance. Interior regions are uniform across urban and rural. Coastal urban runs slightly higher than coastal rural. The islands show the widest within-category spread: rural island policies are the most profitable segment in the portfolio at around 65% loss ratio, while urban island policies sit at around 77%. An 11-percentage-point spread within a single municipality category points to a real underwriting gap.

```{r}
#| fig-width: 10
#| fig-height: 5

clean |>
  mutate(
    claims_bin = if_else(total_claims > 0, "Claim", "No claim"),
    geo_segment = case_when(
      municipality_type == "Interior" & circulation_area == "Rural" ~ "Int·Rur",
      municipality_type == "Interior" & circulation_area == "Urban" ~ "Int·Urb",
      municipality_type == "Coastal"  & circulation_area == "Rural" ~ "Coa·Rur",
      municipality_type == "Coastal"  & circulation_area == "Urban" ~ "Coa·Urb",
      municipality_type == "Islands"  & circulation_area == "Rural" ~ "Isl·Rur",
      municipality_type == "Islands"  & circulation_area == "Urban" ~ "Isl·Urb"
    ),
    geo_segment = factor(geo_segment,
                         levels = c("Coa·Rur","Coa·Urb","Int·Rur",
                                    "Int·Urb","Isl·Rur","Isl·Urb"))
  ) |>
  ggbarstats(
    x = claims_bin,
    y = geo_segment,
    label = "percentage",
    bf.message = FALSE,
    ggtheme = theme_minimal()
  ) +
  scale_fill_manual(values = c("Claim" = house_highlight,
                                "No claim" = house_muted)) +
  labs(x = "Geographic segment  (Int = Interior, Coa = Coastal, Isl = Islands)",
       y = "Share of policies")
```

A chi-square test of independence on the claim-versus-no-claim binary across the six geographic segments rejects the null of independence. The effect size (Cramér's V) is small in absolute terms, which is expected when most policies have no claims, but the differences across segments are statistically robust given the sample size.

The 11-point spread within the islands category indicates that aggregate municipality-level summaries mask underlying within-region heterogeneity.


## Driver age and vehicle age: a risk surface

Driver age and vehicle age are continuous variables. Examining them jointly via the loss-ratio surface reveals which combinations carry the highest expected losses, beyond what either variable alone would suggest.


```{r}
age_interaction <- clean |>
  group_by(driver_age_band, vehicle_age_band) |>
  summarise(
    n_pol = n(),
    premium = sum(total_premium),
    incurred = sum(total_incurred),
    loss_ratio = incurred / premium,
    .groups = "drop"
  ) |>
  filter(n_pol >= 500)

worst_cell <- age_interaction |>
  slice_max(loss_ratio, n = 1)

ggplot(age_interaction,
       aes(x = driver_age_band, y = vehicle_age_band, fill = loss_ratio)) +
  geom_tile(colour = "white", linewidth = 1.5) +
  geom_tile(data = worst_cell,
            colour = "grey15", linewidth = 1.8, fill = NA) +
  geom_text(aes(label = paste0(percent(loss_ratio, accuracy = 0.1),
                                "\nn = ", comma(n_pol))),
            size = 3, lineheight = 0.95, colour = "grey20") +
  scale_fill_distiller(palette = house_sequential, direction = -1,
                       labels = label_percent(accuracy = 1),
                       name = "Loss ratio") +
  scale_x_discrete(expand = c(0, 0)) +
  scale_y_discrete(expand = c(0, 0)) +
  labs(
    title = "Young drivers in older vehicles carry the highest loss ratios",
    subtitle = "Loss ratio by driver age band and vehicle age band; cells with fewer than 500 policies suppressed",
    x = "Driver age band",
    y = "Vehicle age band",
    caption = "Source: Spanish motor insurance portfolio (Mendeley Data).\nCell labels show loss ratio and policy count; the worst cell is outlined."
  ) +
  theme_house(base_size = 11) +
  theme(panel.grid = element_blank())
```
The banded heatmap makes the interaction structure explicit. The under-25 driver band paired with vehicle ages 11 and above shows the highest loss ratios in the surface. The 25 to 49 driver bands show flatter loss-ratio profiles across vehicle ages, indicating that the interaction effect is concentrated in the youngest driver segment. The 65+ band shows a mild upward drift in newer vehicles, consistent with senior drivers in higher-value vehicles carrying distinct loss profiles.


```{r}
clean |>
  filter(total_claims > 0,
         is.finite(loss_ratio),
         loss_ratio >= 0.05, loss_ratio <= 50) |>
  ggbetweenstats(
    x = driver_age_band,
    y = loss_ratio,
    type = "np",
    pairwise.display = "significant",
    p.adjust.method = "fdr",
    plot.type = "boxviolin",
    bf.message = FALSE,
    ggtheme = theme_minimal()
  ) +
  scale_colour_manual(values = c("<25" = house_highlight,
                                  "25-34" = "grey60",
                                  "35-49" = "grey60",
                                  "50-64" = "grey60",
                                  "65+" = "grey60"),
                      guide = "none") +
  scale_fill_manual(values = c("<25" = house_highlight,
                                "25-34" = "grey60",
                                "35-49" = "grey60",
                                "50-64" = "grey60",
                                "65+" = "grey60"),
                    guide = "none") +
  scale_y_continuous(
    trans  = "log10",
    breaks = c(0.1, 0.5, 1, 5, 20),
    labels = c("10%", "50%", "100%", "500%", "2,000%")
  ) +
  coord_cartesian(ylim = c(0.05, 50)) +
  labs(x = "Driver age band",
       y = "Policy-level loss ratio (log scale, claims-positive only)")
```

The Kruskal-Wallis test across driver-age bands rejects the equality null. Pairwise comparisons identify the under-25 segment as significantly different from the 35-49 and 50-64 bands. Both age extremes carry elevated within-segment loss variance, consistent with the wider risk distributions characteristic of younger and older drivers.

## Volatility view

Profitability and volatility are conceptually distinct. A segment can have a low mean loss ratio with high variance, making it unpredictable even if it averages out acceptably. The coefficient of variation of policy-level loss ratio captures this directly.

Policies with non-finite loss ratios (caused by zero or missing premium values) are removed before computing volatility, to avoid distortion from undefined ratios.


```{r}
rvol <- clean |>
  filter(total_claims > 0,
         is.finite(loss_ratio),
         !is.na(loss_ratio)) |>
  group_by(policy_type) |>
  summarise(
    n_pol = n(),
    sd_lr = sd(loss_ratio, na.rm = TRUE),
    .groups = "drop"
  )

vol <- pt |>
  select(policy_type, n_pol, loss_ratio) |>
  rename(segment_lr = loss_ratio) |>
  left_join(rvol |> select(policy_type, sd_lr), by = "policy_type") |>
  mutate(cv_lr = sd_lr / segment_lr)

vol |>
  kable(
    caption = "Segment-level loss ratio and within-segment volatility",
    digits = c(0, 0, 4, 3, 3),
    format.args = list(big.mark = ",")
  )
```

```{r}
#| fig-height: 5

vol <- vol |>
  mutate(highlight = if_else(policy_type == "COMP_N", "COMP_N", "Other"))

ggplot(vol, aes(x = segment_lr, y = cv_lr)) +
  geom_hline(yintercept = mean(vol$cv_lr),
             linetype = "dashed", colour = house_benchmark) +
  geom_vline(xintercept = 0.70,
             linetype = "dashed", colour = house_benchmark) +
  geom_point(aes(size = n_pol, colour = highlight), alpha = 0.85) +
  ggrepel::geom_text_repel(aes(label = policy_type, colour = highlight),
                           size = 3.8, fontface = "bold",
                           point.padding = 0.8, box.padding = 0.6,
                           min.segment.length = 0, seed = 1,
                           show.legend = FALSE) +
  annotate("text", x = 0.71, y = 5,
           label = "70% benchmark", hjust = 0, size = 3,
           colour = house_benchmark, fontface = "italic") +
  annotate("text", x = 0.52, y = mean(vol$cv_lr) + 4,
           label = "Portfolio mean CV", hjust = 0, size = 3,
           colour = house_benchmark, fontface = "italic") +
  scale_x_continuous(labels = label_percent(accuracy = 1),
                     limits = c(0.5, 1.0)) +
  scale_y_continuous(limits = c(0, 140),
                     expand = expansion(mult = c(0, 0.05))) +
  scale_size(range = c(4, 14), guide = "none") +
  scale_colour_manual(values = c("COMP_N" = house_highlight,
                                  "Other" = house_benchmark),
                      guide = "none") +
  labs(
    title = "COMP_N is the worst-performing tier but the most predictable",
    subtitle = "Segment loss ratio versus within-segment volatility, bubble area proportional to policy count",
    x = "Segment loss ratio (profitability axis)",
    y = "Coefficient of variation (volatility axis)",
    caption = "Source: Spanish motor insurance portfolio (Mendeley Data).\nBubble size: number of policies in segment. Reference lines: 70% loss-ratio benchmark and portfolio-mean CV."
  ) +
  theme_house(base_size = 11)
```


The two-axis view separates profitability from predictability. COMP_N has the worst segment loss ratio at 93.5% but the lowest within-segment volatility (CV 7.6), meaning its loss experience is unprofitable but highly predictable. COMP_E shows the opposite profile: a moderately high mean loss ratio of 73.9% paired with the highest volatility in the portfolio (CV 123), driven by occasional large-severity claims that dominate the segment's variance. CC combines a moderate segment loss ratio of 63.5% with mid-range volatility, the most balanced profile in the book.

These two segments illustrate that profitability and predictability operate on independent dimensions. COMP_N is unprofitable but predictable, while COMP_E is closer to break-even on average and its tail risk dominates the within-segment variance.


## Section summary

The bivariate analysis identifies four loss drivers:

1.  **Temporal deterioration**. The 2022 to 2024 loss-ratio trajectory shows a statistically validated worsening, with severity contributing more than frequency.
2.  **Coverage tier mispricing**. COMP_N runs a loss ratio near 94%, materially worse than every other tier and statistically distinct in pairwise testing.
3.  **Bonus-malus paradox**. The Neutral category carries the highest loss ratio in the portfolio despite middle-of-the-road claim frequency, suggesting either default-tier under-loading or downward-rigid bonus-malus transitions. Section 5 examines the transition dynamics directly.
4.  **Geographic and age-driven concentration**. Urban island policies and young-driver-old-vehicle combinations both show elevated loss intensity, with the latter representing a joint pattern that additive treatment of the two age dimensions would miss.

# Beyond standard techniques

The methods applied so far are drawn from the visual analytics curriculum covered in Lessons 1 through 5. This section adds two methods that sit outside that curriculum, selected because they leverage structural features of the data the standard techniques cannot reach. The first method visualises how policyholders move between bonus-malus categories over time, using the dataset's panel structure. The second method shows how unevenly losses are distributed across the policy population, using a concentration metric borrowed from actuarial practice.

## Bonus-malus transitions over time

The same policyholder can appear in 2022, 2023, and 2024 records. This creates a panel structure that single-year cross-sectional analysis discards. The bonus-malus paradox identified in Section 4 (Neutral carrying higher loss ratio than Bad) raised the question of whether bonus-score transitions are downward-rigid. A Sankey diagram of category transitions answers this directly.

```{r}
if (!requireNamespace("ggalluvial", quietly = TRUE)) {
  install.packages("ggalluvial")
}
library(ggalluvial)
```

```{r}
#| fig-height: 4.5

trans_22_24 <- clean |>
  select(insured_id, year, bonus_score) |>
  pivot_wider(names_from = year, values_from = bonus_score,
              names_prefix = "y") |>
  filter(!is.na(y2022), !is.na(y2024)) |>
  filter(!(y2022 == "Good" & y2024 == "Good")) |>
  count(y2022, y2024, name = "policies") |>
  mutate(across(starts_with("y"), ~ factor(.x, levels = c("Good","Neutral","Bad"))))

ggplot(trans_22_24,
       aes(axis1 = y2022, axis2 = y2024, y = policies)) +
  geom_alluvium(aes(fill = y2022), alpha = 0.75, width = 0.3,
                colour = "white", linewidth = 0.2) +
  geom_stratum(width = 0.3, fill = "grey90", colour = "grey50") +
  geom_text(stat = "stratum", aes(label = after_stat(stratum)),
            size = 3.6, fontface = "bold") +
  scale_x_discrete(limits = c("2022", "2024"), expand = c(0.1, 0.1)) +
  scale_fill_manual(values = c("Good" = "#4C72B0",
                                "Neutral" = "#DDB237",
                                "Bad" = "#C44E52"),
                    name = "2022 starting category") +
  scale_y_continuous(labels = comma) +
  labs(
    title = "Among policyholders who moved at all, most started in Neutral or Bad",
    subtitle = "Bonus-malus transitions 2022 to 2024 (stable-Good policyholders excluded)",
    x = NULL,
    y = "Policyholders",
    caption = "Source: Spanish motor insurance portfolio (Mendeley Data).\nPolicyholders observed in both 2022 and 2024; stable-Good excluded."
  ) +
  theme_house(base_size = 11) +
  theme(legend.position = "top")
```

The Sankey reveals the structure of bonus-malus dynamics among policyholders whose category changed at least once between 2022 and 2024. (Policyholders who remained Good throughout, the largest single group at around 99 percent of the panel, are excluded to surface the transition dynamics.) Among the remaining policyholders, upward movement dominates: most who started in Neutral improved to Good, and most who started in Bad improved to Neutral or Good. A smaller but visible flow shows Good policyholders who moved downward into Neutral or Bad. The mechanism therefore favours recovery rather than long-term entrapment, but it also does not protect Good policyholders from downward drift entirely.

```{r}
trans_summary <- clean |>
  select(insured_id, year, bonus_score) |>
  pivot_wider(names_from = year, values_from = bonus_score,
              names_prefix = "y") |>
  filter(!is.na(y2022), !is.na(y2024)) |>
  mutate(
    move = case_when(
      y2022 == y2024 ~ "Stayed",
      as.integer(y2022) < as.integer(y2024) ~ "Worsened",
      TRUE ~ "Improved"
    )
  ) |>
  count(y2022, move) |>
  group_by(y2022) |>
  mutate(pct = n / sum(n)) |>
  ungroup()

trans_summary |>
  kable(
    caption = "2022 to 2024 bonus-malus transitions, by starting category",
    digits = c(0, 0, 0, 4),
    format.args = list(big.mark = ",")
  )
```

The transition table clarifies the pattern. Good policyholders are highly persistent, with 99.6% remaining Good between 2022 and 2024. Neutral and Bad policyholders mostly improve over the same period: 86.0% of Neutral policyholders improve, while 78.8% of Bad policyholders improve. The bonus-malus mechanism therefore appears to allow recovery from worse categories, rather than trapping policyholders there. The pricing concern surfaced in Section 4.4 is the short-term loss-ratio behaviour of policyholders while they sit in the Neutral and Bad pools, rather than any structural transition rigidity.


## Lorenz curve of loss concentration

A second question worth asking of the portfolio is how unevenly losses are distributed across the population of policies. In a perfectly homogeneous book, every policy would contribute the same share of losses. In a heterogeneous book, a small fraction of policies accounts for the bulk of incurred losses, which is what underwriters mean when they refer to a "concentration of losses." The Lorenz curve visualises this.

```{r}
lor <- clean |>
  arrange(total_incurred) |>
  mutate(
    cum_policies = row_number() / n(),
    cum_incurred = cumsum(total_incurred) / sum(total_incurred)
  )

gini_value <- 1 - 2 * sum((lor$cum_incurred + lag(lor$cum_incurred, default = 0)) / 2) / nrow(lor)
gini_value <- round(gini_value, 3)
```




```{r}
#| fig-height: 5

lor <- clean |>
  arrange(total_incurred) |>
  mutate(
    cum_pol = row_number() / n(),
    cum_incurred = cumsum(total_incurred) / sum(total_incurred)
  )

gini_value <- 1 - 2 * sum((lor$cum_pol[-1] - lor$cum_pol[-nrow(lor)]) *
                            (lor$cum_incurred[-1] + lor$cum_incurred[-nrow(lor)]) / 2)

p90_loss_share <- 1 - lor$cum_incurred[which.min(abs(lor$cum_pol - 0.9))]

ggplot(lor, aes(x = cum_pol, y = cum_incurred)) +
  geom_abline(slope = 1, intercept = 0,
              linetype = "dashed", colour = house_benchmark) +
  geom_ribbon(aes(ymin = cum_incurred, ymax = cum_pol),
              fill = house_highlight, alpha = 0.15) +
  geom_line(colour = house_highlight, linewidth = 1.1) +
  annotate("text",
           x = 0.55, y = 0.62,
           label = "Equity line\n(perfect equality)",
           hjust = 0, size = 3, colour = "grey45",
           fontface = "italic", lineheight = 0.95) +
  annotate("text",
           x = 0.35, y = 0.05,
           label = paste0("Gini = ", round(gini_value, 3)),
           hjust = 0, size = 5, colour = house_highlight,
           fontface = "bold") +
  annotate("curve",
           x = 0.78, xend = 0.93,
           y = 0.18, yend = 0.28,
           arrow = arrow(length = unit(2, "mm"), type = "closed"),
           curvature = -0.3, colour = house_highlight, linewidth = 0.4) +
  annotate("text",
           x = 0.77, y = 0.14,
           label = paste0("Top 10% of policyholders\nbear ",
                          round(p90_loss_share * 100),
                          "% of incurred losses"),
           hjust = 1, size = 3.1, colour = house_highlight, lineheight = 0.95) +
  scale_x_continuous(labels = label_percent(accuracy = 1),
                     expand = c(0, 0)) +
  scale_y_continuous(labels = label_percent(accuracy = 1),
                     expand = c(0, 0)) +
  labs(
    title = paste0("Top 10% of policyholders bear ",
                   round(p90_loss_share * 100),
                   "% of incurred losses"),
    subtitle = paste0("Lorenz curve of cumulative incurred losses (Gini = ",
                      round(gini_value, 3), ")"),
    x = "Cumulative share of policyholders (sorted by incurred)",
    y = "Cumulative share of incurred losses",
    caption = "Source: Spanish motor insurance portfolio (Mendeley Data)."
  ) +
  theme_house(base_size = 11)
```

The interpretation is direct. The dashed diagonal represents perfect equality, where the bottom X% of policies would account for X% of losses. The red curve shows reality. The bottom 85% of policies (those with no claims) contribute zero losses. The top 5% of policies account for the majority of total incurred. The Gini coefficient is approximately 0.956, indicating extreme concentration.

```{r}
decile_view <- clean |>
  arrange(total_incurred) |>
  mutate(decile = ntile(total_incurred, 10)) |>
  group_by(decile) |>
  summarise(
    n_pol = n(),
    share_premium = sum(total_premium) / sum(clean$total_premium),
    share_incurred = sum(total_incurred) / sum(clean$total_incurred),
    .groups = "drop"
  )

decile_long <- decile_view |>
  pivot_longer(c(share_premium, share_incurred),
               names_to = "metric", values_to = "share") |>
  mutate(metric = factor(metric,
                          levels = c("share_premium", "share_incurred"),
                          labels = c("Share of premium",
                                     "Share of incurred")))

top10_premium <- decile_view$share_premium[10]
top10_incurred <- decile_view$share_incurred[10]

ggplot(decile_long, aes(x = factor(decile), y = share, fill = metric)) +
  geom_col(position = position_dodge(width = 0.8), width = 0.75) +
  geom_text(data = decile_long |> filter(decile == 10),
            aes(label = percent(share, accuracy = 1)),
            position = position_dodge(width = 0.8),
            vjust = -0.5, size = 3.2, fontface = "bold",
            colour = c(house_benchmark, house_highlight)) +
  scale_y_continuous(labels = label_percent(accuracy = 1),
                     expand = expansion(mult = c(0, 0.12))) +
  scale_fill_manual(values = c("Share of premium"  = house_benchmark,
                                "Share of incurred" = house_highlight),
                    name = NULL) +
  labs(
    title = paste0("The top decile contributes ",
                   percent(top10_premium, accuracy = 1),
                   " of premium but ",
                   percent(top10_incurred, accuracy = 1),
                   " of incurred losses"),
    subtitle = "Premium and incurred-loss shares by decile of policies (ranked low to high by incurred losses)",
    x = "Decile of policies",
    y = "Share of portfolio total",
    caption = "Source: Spanish motor insurance portfolio (Mendeley Data)."
  ) +
  theme_house(base_size = 11) +
  theme(legend.position = "top")
```

The decile view sharpens the Lorenz finding. The top decile of policies accounts for the overwhelming share of incurred losses. The premium share contributed by the same decile is significantly smaller. The two-axis gap shows that premium charged does not currently scale with incurred losses, which is the empirical manifestation of the segment-level mismatches diagnosed in Section 4.


## Cluster heatmap of segment risk profiles

Segments examined one variable at a time can mask the multivariate structure of risk. Two segments may share a moderate loss ratio while behaving very differently on frequency, severity, and premium adequacy. A cluster heatmap addresses this by computing a multi-metric profile for each segment, standardising the metrics to a common scale, and using hierarchical clustering to group segments that share similar profiles. The dendrogram makes the clustering visible.

```{r}
if (!requireNamespace("heatmaply", quietly = TRUE)) {
  install.packages("heatmaply")
}
library(heatmaply)
```

```{r}
#| fig-height: 6.5

seg_profile <- clean |>
  filter(total_premium > 0, total_exposure > 0) |>
  group_by(policy_type, bonus_score) |>
  summarise(
    n_pol      = n(),
    loss_ratio = sum(total_incurred) / sum(total_premium),
    frequency  = sum(total_claims) / sum(total_exposure),
    severity   = if_else(sum(total_claims) > 0,
                         sum(total_incurred) / sum(total_claims),
                         NA_real_),
    avg_prem   = sum(total_premium) / sum(total_exposure),
    .groups = "drop"
  ) |>
  filter(n_pol >= 100, !is.na(severity)) |>
  mutate(segment = paste(policy_type, bonus_score, sep = " · ")) |>
  select(segment, loss_ratio, frequency, severity, avg_prem)

seg_matrix <- seg_profile |>
  column_to_rownames("segment") |>
  scale() |>
  as.matrix()

heatmaply(
  seg_matrix,
  colors = colorRampPalette(c("#4C72B0", "white", house_highlight))(50),
  dendrogram = "both",
  k_row = 3,
  fontsize_row = 9,
  fontsize_col = 10,
  column_text_angle = 0,
  main = "Segments cluster into three distinct risk profiles",
  xlab = NULL,
  ylab = NULL,
  margins = c(60, 120, 50, 20),
  plot_method = "ggplot"
)
```

The clustering algorithm groups segments by similarity across all four standardised metrics simultaneously. Three groups emerge. The first group contains all three COMP_N combinations (Good, Neutral, Bad) together, defined by elevated claim frequency and elevated average premium relative to the portfolio mean. That all three bonus-malus tiers within COMP_N cluster together tells us the mispricing identified in Section 4.3 is structural to the product rather than driven by customer mix. The second and largest group spans the CC, COMP_E, TP, and TPG segments in Good and Neutral bonus tiers, exhibiting balanced profiles close to the portfolio mean. The third group is a singleton: TPG · Bad, isolated from every other segment by its very high loss ratio and very high severity. With only a few hundred policies, this segment is small in volume but statistically distinct, and warrants individual underwriting attention rather than being averaged into the broader TPG tariff. The dendrogram quantifies these separations: the largest vertical jump isolates the TPG · Bad singleton, with COMP_N's three-segment cluster separating from the main group at the next level down.


# Summary

This analysis of a Spanish motor insurance portfolio across 2022 to 2024 examined the distributional shape of policy, vehicle, driver, and geographic characteristics, and investigated bivariate relationships with portfolio profitability and volatility. The key findings are summarised below.

## Key findings

- The portfolio loss ratio rose from 65.7 percent in 2022 to 74.7 percent in 2024. The deterioration is statistically validated through non-overlapping bootstrap confidence intervals and is driven by claim severity rather than claim frequency.
- Loss ratio varies monotonically across coverage tiers, from 54 percent in basic third-party to approximately 94 percent in comprehensive-without-excess (COMP_N). Pairwise non-parametric tests confirm that COMP_N is statistically distinct from every other tier.
- The bonus-malus categories show an unexpected pattern: the Neutral tier carries a higher loss ratio than the Bad tier despite collecting less premium per exposure. Frequency rises monotonically across the three categories as expected, but the loss-ratio ordering inverts at the top.
- Geographic loss intensity differs sharply within municipality categories. Urban island policies run 11 percentage points above rural island policies, the widest within-category spread in the geographic dimension.
- Driver age and vehicle age interact non-additively in their relationship with loss ratio. The under-25 driver band paired with older vehicles shows the highest loss intensity in the joint surface.
- Coverage tier and bonus-malus also interact in the multivariate clustering: all three COMP_N combinations group together regardless of bonus-malus status, indicating that the elevated loss profile is structural to the product rather than driven by customer mix.
- Loss concentration is extreme. The Lorenz curve of incurred losses yields a Gini coefficient of approximately 0.96, with the top decile of policies bearing approximately 99 percent of total incurred losses.

## Recommendations for future research

- **Confirmatory analysis.** This study is exploratory. Future work should apply formal causal-inference methods to confirm whether the segment-level loss-ratio differences reflect underlying risk heterogeneity or unobserved confounding.
- **Additional variables.** The Mendeley dataset is rich on policy and vehicle characteristics but contains no driving-behaviour telematics, no claim-type detail beyond seven coverage categories, and no geographic granularity below the municipality type. Incorporating these would refine the segment-level diagnostics.
- **Longitudinal modelling.** The panel structure used in the Sankey analysis could be extended to a full longitudinal model of policyholder transitions, including survival analysis of policy cancellation and Markov chain modelling of bonus-malus dynamics.
- **External benchmarking.** Comparing the observed loss ratios and segment patterns against published Spanish motor market benchmarks would clarify whether the findings are portfolio-specific or generalisable.

## Methodological reflection

The exercise demonstrates that visual analytics serves three distinct functions in a portfolio review. The first is descriptive, characterising the shape and concentration of the variables that constitute the book. The second is diagnostic, surfacing the segments where premium and exposure are out of balance. The third is communicative, presenting findings in a form that supports further investigation. The bonus methods applied in Section 5 illustrate that the most informative approaches sometimes exploit data-structural features (panel observation, concentration of loss, multivariate clustering) rather than summarising marginal distributions.



# References

-   Swiss Re Institute (2019). *sigma 4/2019: Advanced analytics: unlocking new frontiers in P&C insurance.* Zurich: Swiss Re Management Ltd.
-   Mendeley Data. *Spanish motor vehicle insurance portfolio dataset.*
-   Wickham, H. et al. (2019). Welcome to the Tidyverse. *Journal of Open Source Software*, 4(43), 1686.
-   Patil, I. (2021). Visualizations with statistical details: The ggstatsplot approach. *Journal of Open Source Software*, 6(61), 3167.
-   Kay, M. (2024). ggdist: Visualizations of Distributions and Uncertainty. *R package version 3.3.2*.
-   Brunson, J. C. (2020). ggalluvial: Layered Grammar for Alluvial Plots. *Journal of Open Source Software*, 5(49), 2017.
