---
title: "Take Home Exercise 2 - VAST Challenge MC 1"
subtitle: "Reconstructing the TenantThread Embargo Breach"
author: "Yap Jit Wei"
date: "May 26, 2026"
date-modified: "last-modified"
format:
  html:
    toc: true
    toc-depth: 4
    number-sections: true
    code-fold: true
    code-summary: "Show the code"

execute:
  warning: false
  message: false
  freeze: auto
---

# Overview

## The Case

On 5 June 2046 at 6:00 PM, the **Project HarborCrest** merger between
**TenantThread** and **CivicLoom** was scheduled to be publicly announced.
At around 5:00 PM that afternoon, an hour before the embargo lifted,
details of the deal began appearing on the **TenantThread** Flex account.
The embargo had been broken. **CivicLoom**'s legal team now needs to
understand what happened.

This is the backdrop. **TenantThread** sells property management software
to landlords, and one of its products scores tenants based on payment
timing, maintenance requests, and message tone. A local newspaper has
been running an exposé on these practices for weeks, and the hashtag
**#AlgorithmicEviction** has been gaining traction. The merger with
**CivicLoom** was meant to bring in capital and rebrand the platform.

Seven AI agents run **TenantThread**'s corporate communications. They
cover legal, PR, social media, trust and safety, two intern roles,
and a compliance officer known as the **Judge**, which was installed
specifically to keep embargoed information from being released. The
question this report sets out to answer is:

> **Was the merger leaked on purpose, or did the system fail under
> pressure?**

## Answer in Brief

::: callout-important
## The short version

Neither framing in the brief fits cleanly. Three things are true
at once:

1. **The agents probably did not cause the original merger leak.**
The original information escape came from sources the team
themselves flagged as external — a **CivicLoom** board member
leak, an employee Slack screenshot, and anonymous accounts using
**#CivicLoom #6PM** that **Platform Trust** explicitly called
"not ours" in an internal message at 1 PM.

2. **The agents did breach the spirit of the embargo.** At 5 PM,
once **SaltWind** published, **Legal** and **Social Manager**
executed a pre-staged two-track response across personal and
anonymous channels while **Official Flex** stayed silent. The
counter-narrative was ready to deploy the moment **SaltWind**
gave it cover.

3. **The compliance architecture failed at the point that
mattered.** The **Judge** was watching aggregation exposure climb
all afternoon and issued a COMPLIANCE_WARNING at 3 PM. It then
sent zero messages between 4 PM and 6 PM, and its scope never
covered the personal and anonymous channels where the response
moved through anyway.

The detailed reconstruction follows in Sections 5 through 8.

:::

## The Data

The dataset is a single JSON file covering 23 rounds spanning
17 May 2046 to 5 June 2046. The structure has two phases.
Rounds 1 to 13 are sampled once per business day across the
two-week buildup. Rounds 14 to 23 are sampled once per hour
on a single day, 5 June, from 9 AM to 6 PM. The hourly
sampling on the crisis day means a single afternoon gets the
same number of observations as the two weeks of lead-up. This
lets the analysis follow the breach minute by minute as it
happens.

Within each round there are three main blocks:

1. Environment context, which includes the event narrative,
   headline, market snapshot, media events, social state,
   external actor actions, agents unavailable, and critical
   deadlines.
2. Communications, which contains every message sent by any agent
   during the round. Each message carries a channel, recipients,
   message type, and for a smaller subset, the agent's internal
   thoughts at the moment of sending (reacting, rationalizing,
   and deliberating).
3. Participants, which gives agent-level metadata for the round.

In total there are 912 messages across 6 channels (comms_huddle,
one_on_one_chat, side_huddle, official_post, personal_post,
anonymous_post). Only 86 of these messages carry populated
internal-state text, so any chart that relies on internal thoughts
will need to handle missing values explicitly.

## The Three Tasks

The VAST Challenge sets three tasks for this mini-challenge. The
later sections of this report are organised around them so each
task is answered in its own chapter.

**Task 1: Reconstruct the sequence of events.** What were the key
actions, causal relationships, and decision points that led to the
embargoed information being released? Which system elements allowed
the post to get past the **Judge**?

**Task 2: Compare normal behavior to the leak.** The evasion of
the embargo was new. What did typical agent behavior look like
before the leak, and how does the behavior that led to the release
differ from it?

**Task 3: Find leading indicators.** Were there prior occasions
where an agent's actual behavior differed from its expected
behavior? Were there earlier near-misses that looked like the
leak-leading behavior? And if so, why did nothing happen?

## Analytical Approach

The investigation decomposes the brief's three tasks into three
analytical questions, each tied to a specific method and section
of the report.

```{mermaid}
%%| label: fig-investigation-framework
%%| fig-cap: "Investigation framework: each task is decomposed into a question, a method, and a target section."

flowchart LR
    Q1["<b>Question 1</b><br/>What happened<br/>during the breach?"]
    Q2["<b>Question 2</b><br/>How does breach behaviour<br/>differ from normal?"]
    Q3["<b>Question 3</b><br/>What earlier signals<br/>could have warned us?"]

    M1["<b>Method</b><br/>Hour-by-hour timeline<br/>+ post-sequence analysis<br/>+ Judge presence check"]
    M2["<b>Method</b><br/>Baseline (rounds 1–13)<br/>vs crisis day (rounds 14–23)<br/>per-agent rate comparison"]
    M3["<b>Method</b><br/>Near-miss audit<br/>+ anonymous-channel scan<br/>+ Judge scope vs breach"]

    S1["<b>Section 5</b><br/>Task 1 answer"]
    S2["<b>Section 6</b><br/>Task 2 answer"]
    S3["<b>Section 7</b><br/>Task 3 answer"]

    Q1 --> M1 --> S1
    Q2 --> M2 --> S2
    Q3 --> M3 --> S3

    classDef question fill:#FDECEC,stroke:#C44E52,stroke-width:2px,color:#000
    classDef method fill:#F5F5F5,stroke:#666,stroke-width:1.5px,color:#000
    classDef section fill:#FFF4E6,stroke:#C44E52,stroke-width:1.5px,color:#000

    class Q1,Q2,Q3 question
    class M1,M2,M3 method
    class S1,S2,S3 section
```

Section 4 (Exploratory Analysis) precedes the three task answers
and establishes the communication landscape used as context
throughout. Section 8 reconciles the three answers into a single
verdict. This way of working follows the patterns-of-life framing
rewarded in past VAST Challenge submissions (Guan, 2017).

# Setup

## Loading Packages

```{r}
pacman::p_load(
  # JSON and tidy data
  jsonlite, tidyverse, lubridate,
  # Visualisation
  ggplot2,
  # Network analysis
  tidygraph, ggraph,
  # Tables
  knitr, DT
)
```

The visual stack stays within the ggplot2 family, including the network view (`ggraph` rather than `visNetwork`). Tables use `knitr::kable()` for static sanity-check tables in Section 3 and `DT::datatable()` for the interactive Round 22 post inventory in Section 5.2.

## Loading Data

The MC1 JSON is loaded with simplifyVector set to FALSE so the
nested list structure is kept intact. This matters because the
internal_state field is null in about 90 percent of messages,
and forcing simplification fails badly when nested fields are
only partly populated.

```{r}
mc1_raw <- fromJSON("Data/MC1_final_00.json", simplifyVector = FALSE)
length(mc1_raw$rounds)
```

The result should be 23 rounds, covering the period from the
start of the investigation window to the moment the embargo
lifts.

## Defining Common Styles

A consistent visual language across the report makes the
analysis easier to follow. The design philosophy below is the
same one used in Take-home Exercise 1, applied to a new dataset.

Four principles guide the chart styling.

The first is one focal colour per chart. A single red marks the
actor, round, or category being discussed in the surrounding
prose. Every other element is muted grey so the reader's eye
goes to the right place without ambiguity.

The second is reference values in muted grey. Benchmarks such
as the embargo line, the baseline mean, or a Judge threshold
sit in a mid-grey so they recede behind the data.

The third is sequential palette only where direction matters.
The channel field is genuinely ordered from private to public,
so a grey-to-red scale is used to encode that order. Where the
variable has no inherent order, such as agent identity, facets
or labels are used instead of colour.

The fourth is finding-led titles. Chart titles state the
headline finding rather than describing the chart type. The
subtitle names the underlying variables and the population.

The colour constants are factored out of the theme so they can
be referenced by name in any chart. This keeps the palette
consistent across the report and makes any future palette
change a one-line edit.

```{r}
theme_house <- function(base_size = 11) {
  theme_minimal(base_size = base_size) +
    theme(
      plot.title = element_text(face = "bold", size = base_size + 2),
      plot.subtitle = element_text(colour = "grey40",
                                   size = base_size - 1,
                                   margin = margin(b = 10)),
      plot.caption = element_text(colour = "grey50",
                                  size = base_size - 3,
                                  hjust = 0),
      panel.grid.major.x = element_blank(),
      panel.grid.minor = element_blank(),
      axis.text = element_text(size = base_size - 2),
      axis.title = element_text(size = base_size - 1.5),
      axis.title.x = element_text(margin = margin(t = 8)),
      axis.title.y = element_text(margin = margin(r = 8)),
      plot.margin = margin(t = 10, r = 15, b = 15, l = 10),
      legend.position = "top",
      legend.title = element_text(size = base_size - 2, face = "bold"),
      legend.text = element_text(size = base_size - 2)
    )
}

house_highlight  <- "#C44E52"
house_muted      <- "grey75"
house_benchmark  <- "grey40"
house_sequential <- "RdYlBu"
```

A small set of lookups is still needed for axis labels, for the
channel ordering, and for reconciling the role strings used in
the recipients column with the canonical agent_id strings used
everywhere else.

The channel palette runs from light grey for the most private
channel up to the focal red for the most public channel. The
two public-post channels (anonymous_post and official_post) are
given red tones because they are the channels through which an
embargo can actually be broken, so any chart using channel as a
fill makes a private-to-public shift visible at a glance.

```{r}
AGENT_LABELS <- c(
  legal_agent          = "Legal",
  quality_agent        = "Platform Trust",
  social_media_agent = "Social Manager",
  pr_agent             = "PR",
  intern_agent         = "Intern",
  pr_intern_agent      = "PR Intern",
  judge_agent     = "Judge"
)

CHANNEL_COLORS <- c(
  comms_huddle    = "grey85",
  side_huddle     = "grey70",
  one_on_one_chat = "grey55",
  personal_post   = "grey40",
  anonymous_post  = "#E89A9C",
  official_post   = house_highlight
)

ROLE_TO_ID <- c(
  legal           = "legal_agent",
  platform_trust  = "quality_agent",
  social_manager  = "social_media_agent",
  social_media    = "social_media_agent",
  pr              = "pr_agent",
  intern          = "intern_agent",
  pr_intern       = "pr_intern_agent",
  judge           = "judge_agent"
)
```

# Data Wrangling

The raw JSON is a deeply nested list. Before any analysis can
happen, it needs to be flattened into a small set of tidy tibbles
that downstream chart code can use. This section builds four
working tables and runs a few sanity checks against them.

The four tables are:

1. `rounds` — one row per hour-round, with the environment
   context flattened into columns. 23 rows.
2. `messages` — one row per communication, with sender, channel,
   recipients, message text, and the three internal-state fields
   exposed as columns. Around 912 rows.
3. `edges` — one row per sender-recipient pair per round, derived
   from `messages` by unnesting the recipients list. Used for
   network plots.
4. `agent_round` — one row per agent per round, aggregating
   message counts and channel mix. Used for baseline behaviour
   charts.

## Inspecting the JSON Structure

A quick look at the top of the structure confirms what fields
exist before any flattening logic is written.

```{r}
str(mc1_raw, max.level = 2)
```

The object has one top-level key, `rounds`, which is a list of
23 elements. Each element has four keys: `hour`,
`environment_context`, `communications`, and `participants`.
The communications list within each round holds the message
records that drive most of this analysis.

A look at a single message confirms the message-level schema:

```{r}
mc1_raw$rounds[[1]]$communications[[1]] %>% str(max.level = 2)
```

Each message has identifiers (`message_id`, `agent_id`,
`agent_role`, `agent_label`), an `internal_state` block with
three sub-fields (`reacting`, `rationalizing`, `deliberating`),
and routing information (`channel`, `recipients`,
`message_type`, `responding_to`, `message_text`).

The recipients field is a list, since a message can have one
recipient, several, or the special value `"ALL"` for broadcasts.
This is the reason for keeping it as a list-column at first and
unnesting it only when the edge list is built.

## Building the Rounds Table

The rounds table pulls one row per hour-round and flattens the
environment_context fields that are scalars. List-valued fields
within the context (such as media_events or critical_deadlines)
are kept as list-columns for now, since they will be unnested
on demand later.

```{r}
safe_chr <- function(x) if (is.null(x)) NA_character_ else as.character(x)

rounds <- tibble(round_data = mc1_raw$rounds) %>%
  mutate(
    hour            = map_chr(round_data, "hour"),
    hour_dt         = ymd_hms(hour),
    round_num       = row_number(),
    headline        = map_chr(round_data,
                              ~ safe_chr(.x$environment_context$event_headline)),
    narrative       = map_chr(round_data,
                              ~ safe_chr(.x$environment_context$event_narrative)),
    n_comms         = map_int(round_data,
                              ~ length(.x$communications)),
    n_participants  = map_int(round_data,
                              ~ length(.x$participants))
  ) %>%
  select(-round_data)

glimpse(rounds)
```

The helper `safe_chr` converts NULLs to NAs explicitly. This is
needed because some rounds have an empty narrative or headline,
and `map_chr` would otherwise fail when it encounters a NULL.

## Building the Messages Table

The messages table is the workhorse of the report. It pulls one
row per communication and exposes every field that downstream
plots will use.

```{r}
messages <- tibble(round_data = mc1_raw$rounds) %>%
  mutate(
    hour  = map_chr(round_data, "hour"),
    comms = map(round_data, "communications")
  ) %>%
  select(hour, comms) %>%
  unnest_longer(comms) %>%
  mutate(
    hour_dt        = ymd_hms(hour),
    message_id     = map_chr(comms, "message_id"),
    agent_id       = map_chr(comms, "agent_id"),
    agent_role     = map_chr(comms, "agent_role"),
    agent_label    = map_chr(comms, "agent_label"),
    channel        = map_chr(comms, "channel"),
    message_type   = map_chr(comms, ~ safe_chr(.x$message_type)),
    responding_to  = map_chr(comms, ~ safe_chr(.x$responding_to)),
    content        = map_chr(comms, ~ safe_chr(.x$content)),
    timestamp      = map_chr(comms, ~ safe_chr(.x$timestamp)),
    recipients     = map(comms, "recipients"),
    reacting       = map_chr(comms,
                             ~ safe_chr(.x$internal_state$reacting)),
    rationalizing  = map_chr(comms,
                             ~ safe_chr(.x$internal_state$rationalizing)),
    deliberating   = map_chr(comms,
                             ~ safe_chr(.x$internal_state$deliberating))
  ) %>%
  left_join(rounds %>% select(hour, round_num), by = "hour") %>%
  select(-comms)

glimpse(messages)
```

The join with `rounds` brings in the round number for each
message, which makes downstream grouping cleaner than parsing
the timestamp every time.

The `recipients` column is kept as a list, since a message can
have one recipient, several, or the value "ALL". Unnesting it
here would inflate the table and make per-message charts
overcount.

## Deriving the Edge List

The edge list is used for network plots. It is created by
unnesting the recipients column, which produces one row per
sender-recipient pair per message. Self-loops (where an agent
appears as both sender and recipient via "ALL") are kept,
because they reflect a real broadcast event rather than a data
artefact.

```{r}
edges <- messages %>%
  select(round_num, hour_dt, message_id, agent_id, channel, recipients) %>%
  unnest_longer(recipients, values_to = "recipient") %>%
  filter(!is.na(recipient)) %>%
  rename(sender = agent_id)

glimpse(edges)
```

A quick look at what values appear in the recipient column
tells us whether the data uses agent IDs, role names, or other
labels for recipients:

```{r}
edges %>% count(recipient, sort = TRUE) %>% head(15) %>% kable()
```

The "ALL" value represents a broadcast to the comms huddle,
which is functionally a one-to-many edge. For network plots
that need clean dyads, ALL can be expanded to all other agents
in that round; for plots that care about volume, it can be
treated as a single edge to a synthetic "ALL" node.

## Building the Agent-Round Summary

The agent-round table is a per-agent, per-round aggregation of
activity. It is the foundation for the baseline-behaviour
charts in Task 2 and the anomaly comparisons across rounds.

```{r}
agent_round <- messages %>%
  group_by(round_num, hour_dt, agent_id, agent_role) %>%
  summarise(
    n_messages       = n(),
    n_with_internal  = sum(!is.na(deliberating) |
                           !is.na(rationalizing) |
                           !is.na(reacting)),
    n_official_post  = sum(channel == "official_post"),
    n_anonymous_post = sum(channel == "anonymous_post"),
    n_personal_post  = sum(channel == "personal_post"),
    n_private        = sum(channel %in% c("comms_huddle",
                                          "side_huddle",
                                          "one_on_one_chat")),
    .groups = "drop"
  )

glimpse(agent_round)
```

This table flattens the channel mix into separate counts, which
makes per-agent stacked bar charts straightforward later. Rounds
where an agent sent no messages will be missing from this table,
which is fine for activity plots but may need to be filled with
zeros for some heatmap views. That fill step is handled in
Section 4 where it is actually used.

## Sanity Checks

A few checks confirm that the wrangling produced what was
expected before the analysis begins.

**Row counts.**

```{r}
list(
  rounds      = nrow(rounds),
  messages    = nrow(messages),
  edges       = nrow(edges),
  agent_round = nrow(agent_round)
) %>% as_tibble() %>% kable()
```

The rounds table should have 23 rows. The messages table
should be close to 912. The edges table will be larger than
the messages table because of recipient unnesting.

**Distinct agents and channels.**

```{r}
messages %>%
  count(agent_id, agent_role, sort = TRUE) %>%
  kable()
```

```{r}
messages %>%
  count(channel, sort = TRUE) %>%
  kable()
```

These two tables confirm that the seven agents and six channels
described in the brief all appear in the data, and surface any
unexpected variants in agent_id strings (for instance,
social_media_agent versus social_manager_agent).

**Coverage of internal_state.**

```{r}
messages %>%
  summarise(
    total                = n(),
    has_reacting         = sum(!is.na(reacting)),
    has_rationalizing    = sum(!is.na(rationalizing)),
    has_deliberating     = sum(!is.na(deliberating)),
    has_any_internal     = sum(!is.na(reacting) |
                               !is.na(rationalizing) |
                               !is.na(deliberating))
  ) %>%
  kable()
```

About 86 of 912 messages carry populated internal-state text.
Any chart that uses these fields will need to filter for
non-NA values rather than assume universal coverage.

**Time range.**

```{r}
range(messages$hour_dt) %>%
  format("%Y-%m-%d %H:%M") %>%
  paste(collapse = " to ") %>%
  cat()
```

The expected range is 17 May 2046 09:00 to 5 June 2046 18:00.

## Assumptions and Interpretive Decisions

::: callout-note
## Three decisions affecting downstream analysis

1. **Broadcasts modelled as edges to a synthetic ALL node.**
The recipients field uses the literal value `"ALL"` for comms
huddle broadcasts. The network in Section 4.3 treats `ALL` as
a single synthetic node rather than expanding each broadcast
into edges to every other agent. This keeps the network legible
but slightly understates the **Judge**'s structural isolation,
since fully expanded broadcasts would emphasise even more that
the **Judge** has no targeted connections.

2. **Empty recipient lists left unjoined.** The `one_on_one_chat`
and `side_huddle` channels contain messages with empty
`recipients` arrays — the recipient is implied by conversational
context rather than listed. These messages are excluded from the
edge list and therefore do not appear in the network plot.
Volume-based charts (Sections 4.1, 5.1, 6) include these messages
in their counts.

3. **Internal-state coverage treated as missing-not-random.**
Only 86 of 912 messages carry populated `internal_state` text.
Where these deliberations are quoted (Section 7.1 **@ElenaMarquez**
incident, Section 7.3 **Judge** COMPLIANCE_WARNING), the report
treats them as illustrative rather than representative. No chart
aggregates over the internal-state field.

:::

# Exploratory Analysis

Before answering any of the three tasks, this section maps the
communication landscape across the full dataset — which agents
post publicly, on which channels, and what the structural shape
of the team's coordination looks like. The charts here use all
23 rounds to give a complete view of the surface, since some of
the most distinctive patterns (**Legal**'s anonymous posting,
the **Judge**'s network position) only emerge against the full
window. The formal baseline-versus-crisis comparison is performed
in Section 6 using only the 13 baseline rounds.


## Activity Volume Over Time

The first view shows how many messages were sent in each round,
split into private agent-to-agent traffic and posts to public
channels. The reader can see both the overall rhythm of the
team and the channel mix that makes up that rhythm.

```{r}
#| label: fig-volume-over-time
#| fig-cap: "Source: MC1_final_00.json"
#| fig-width: 9
#| fig-height: 4.5

vol_by_round <- messages %>%
  mutate(
    channel_group = if_else(
      channel %in% c("official_post", "anonymous_post", "personal_post"),
      "Public posts",
      "Private messages"
    ),
    channel_group = factor(channel_group,
                           levels = c("Private messages", "Public posts"))
  ) %>%
  count(round_num, hour_dt, channel_group)

ggplot(vol_by_round,
       aes(x = round_num, y = n, fill = channel_group)) +
  geom_col(width = 0.8) +
  scale_fill_manual(
    values = c("Private messages" = house_muted,
               "Public posts"     = house_highlight),
    name   = NULL
  ) +
  scale_x_continuous(breaks = seq(1, max(vol_by_round$round_num), by = 2)) +
  labs(
    title    = "Public posting is constant, with the final round reaching its peak",
    subtitle = "Messages per round, split into private agent-to-agent traffic and public posts",
    x        = "Round number",
    y        = "Messages sent",
    caption  = "Source: MC1_final_00.json"
  ) +
  theme_house()
```

::: callout-note
## Insights

- Rounds 1 to 13 each represent a full business day. Message volume sits in a steady 25 to 45 range with a clear lull around the middle of the window.
- Rounds 14 to 23 are hourly snapshots of a single day (5 June 2046). The rightmost ten bars therefore compress a single afternoon into the same horizontal space the baseline uses for two weeks.
- The rising volume in the rightmost bars is the cumulative pressure of a ten-hour breach day building toward the 6 PM embargo lift.

:::

## Channel Mix by Agent

The next view zooms into public-channel activity only. Private
agent-to-agent messages are excluded because they would
otherwise swamp the chart at roughly an order of magnitude
larger than any public-channel volume. Agents who never posted
publicly are not shown. Within each bar, the three
public-channel types are stacked so the channel mix per agent
can be read at a glance.

```{r}
#| label: fig-channel-by-agent
#| fig-cap: "Source: MC1_final_00.json"
#| fig-width: 9
#| fig-height: 4.5

public_by_agent <- messages %>%
  filter(agent_id %in% names(AGENT_LABELS),
         channel  %in% c("official_post", "anonymous_post", "personal_post")) %>%
  mutate(
    channel = factor(channel,
                     levels = c("personal_post",
                                "anonymous_post",
                                "official_post"),
                     labels = c("Personal posts",
                                "Anonymous posts",
                                "Official Flex"))
  ) %>%
  count(agent_id, channel) %>%
  mutate(agent_label = AGENT_LABELS[agent_id])

agent_order <- public_by_agent %>%
  group_by(agent_label) %>%
  summarise(total = sum(n)) %>%
  arrange(total) %>%
  pull(agent_label)

public_by_agent <- public_by_agent %>%
  mutate(agent_label = factor(agent_label, levels = agent_order))

ggplot(public_by_agent,
       aes(x = n, y = agent_label, fill = channel)) +
  geom_col(width = 0.7) +
  scale_fill_manual(
    values = c("Personal posts"  = "grey75",
               "Anonymous posts" = "grey50",
               "Official Flex"   = house_highlight),
    name = NULL
  ) +
  labs(
    title    = "Three agents operate Official Flex across the full window",
    subtitle = "Public-channel post counts by agent across all 23 rounds",
    x        = "Posts sent",
    y        = NULL,
    caption  = "Private agent-to-agent messages excluded. Source: MC1_final_00.json"
  ) +
  theme_house() +
  theme(panel.grid.major.y = element_blank(),
        panel.grid.major.x = element_line(colour = "grey90"))
```

::: callout-note
## Insights

- **Social Manager** is the most active public poster across the full window. 26 public posts in total, split between Personal posts (16) and **Official Flex** (10). The split itself is informative: **Social Manager**'s baseline use of **Official Flex** (9 posts across 13 baseline rounds) gives way to a heavy Personal-post pattern on the crisis day.
- The **Official Flex** account has three operators across the full window: **PR Intern** (12 posts), **Social Manager** (10), and **PR** (6). The leading operator changes between baseline and crisis day. **Social Manager** leads **Official Flex** during the baseline; **PR Intern** takes over during the crisis day. Section 6.3 unpacks this shift.
- All 12 anonymous posts in the dataset are from **Legal**, and they all occur on the crisis day. This is the most distinctive single pattern in the data and is treated in detail in Section 7.2.

:::

## Who Talks to Whom

The third view collapses agent-to-agent communication into a
network. Each node is an agent. Each edge shows the count of
messages between two agents, with thicker edges representing
more traffic. The **Judge** is highlighted in red because the
chart's purpose is to surface its structural isolation against
the team's heavy broadcast coordination.

```{r}
#| label: fig-network-overall
#| fig-width: 9
#| fig-height: 8

edge_summary <- edges %>%
  mutate(recipient = recode(recipient, !!!ROLE_TO_ID),
         recipient = if_else(recipient == "ALL", "ALL_broadcast", recipient)) %>%
  filter((recipient %in% names(AGENT_LABELS)) | recipient == "ALL_broadcast",
         sender %in% names(AGENT_LABELS)) %>%
  count(sender, recipient, name = "weight") %>%
  rename(from = sender, to = recipient)

node_summary <- tibble(
  name = unique(c(edge_summary$from, edge_summary$to))
) %>%
  mutate(
    label = case_when(
      name == "ALL_broadcast" ~ "ALL\n(broadcast)",
      TRUE                    ~ AGENT_LABELS[name]
    ),
    is_focal = name == "judge_agent",
    is_broadcast_node = name == "ALL_broadcast"
  )

g <- tbl_graph(nodes = node_summary,
               edges = edge_summary,
               directed = TRUE)

ggraph(g, layout = "linear", circular = TRUE) +
  geom_edge_arc(aes(width = weight),
                alpha = 0.35,
                colour = "grey50",
                arrow = arrow(length = unit(2.5, "mm"),
                              type = "closed"),
                end_cap = circle(6, "mm"),
                start_cap = circle(6, "mm")) +
  geom_node_point(aes(fill = is_focal,
                      shape = is_broadcast_node),
                  size = 14,
                  colour = "white",
                  stroke = 1.2) +
  geom_node_text(aes(label = label,
                     colour = is_focal),
                 size = 3.5,
                 fontface = "bold",
                 nudge_y = 0.22) +
  scale_edge_width(range = c(0.3, 3),
                   name  = "Messages exchanged") +
  scale_fill_manual(values = c(`FALSE` = house_muted,
                               `TRUE`  = house_highlight),
                    guide  = "none") +
  scale_shape_manual(values = c(`FALSE` = 21,
                                `TRUE`  = 22),
                     guide  = "none") +
  scale_colour_manual(values = c(`FALSE` = "grey30",
                                 `TRUE`  = house_highlight),
                      guide  = "none") +
  coord_cartesian(clip = "off") +
  labs(
    title    = "The Judge has the thinnest network footprint of any agent",
    subtitle = "Directed message flow between agents and the ALL broadcast channel, across all 23 rounds",
    caption  = "Source: MC1_final_00.json"
  ) +
  theme_void(base_size = 11) +
  theme(
    plot.title      = element_text(face = "bold", size = 13),
    plot.subtitle   = element_text(colour = "grey40", size = 10,
                                   margin = margin(b = 10)),
    plot.caption    = element_text(colour = "grey50", size = 8, hjust = 0),
    plot.margin     = margin(t = 10, r = 30, b = 15, l = 30),
    legend.position = "top"
  )
```

::: callout-note
## Insights

- The team coordinates heavily through broadcasts. Most agents have thick edges to the ALL node, with **Legal**, **Platform Trust**, and **Social Manager** each carrying over 100 broadcasts.
- The **Judge** has the thinnest network footprint of any agent. Just 15 broadcasts to the ALL node, one directed message to **Legal**, and one received from **Legal**.
- - For a compliance role designed to flag issues quickly with specific people, that is a structurally thin connection. The behavioural consequence shows up in Section 5.4.


:::

## What This Tells Us About the Communication Surface

The landscape views give a working picture of how the team is built.

::: callout-note
## Insights

- The team's primary coordination channel is the comms huddle broadcast. **Platform Trust**, **Legal**, **Social Manager**, and **PR** each carry between 45 and 100 broadcasts in the baseline window alone. Most coordination happens openly to the full team, not in directed messages.
- Public posting is wider than a single account or single agent. Three agents operate **Official Flex** (**PR Intern**, **Social Manager**, **PR**), and **Legal** is a meaningful anonymous poster on the crisis day. The leak surface is therefore not one channel and not one operator.
- The **Judge** has the thinnest network footprint of any senior agent on the team. Directed messages only exchanged with **Legal**, and the lowest broadcast volume of any agent except the two interns. For a compliance role designed to act quickly on specific issues, that is a thin connection.

:::

The formal comparison between baseline and crisis-day behaviour, agent by agent, is in Section 6.

# Task 1: The Sequence of Events

::: callout-tip
## Task 1 Answer (Summary)

The embargo break on 5 June was a multi-hour cascade. At 5 PM,
**SaltWind** Journal published an exclusive confirming the
**CivicLoom**-**TenantThread** merger one hour ahead of the
official 6 PM announcement. Across the preceding hours, internal
information flow inside **TenantThread** accelerated steadily, with
**Legal** coordinating governance-reform messaging that was
timestamped earlier in the day specifically to be
counter-narrative-ready when the embargo broke.

When **SaltWind** published at 5 PM, **Legal** and **Social Manager**
executed a pre-staged response: a coordinated two-track wave of
personal and anonymous posts. **Legal** handled the legitimacy
narrative, leading every post with merger confirmation and
governance-reform timestamps. **Social Manager** handled the
perception narrative, posting campaign-styled sentiment updates
labelled "EXECUTING" and "WAVE THREE". The **Official Flex**
account, where **PR Intern** operates, stayed silent until 6 PM
when the embargo formally lifted.

The system was worked around through three design gaps. The **Judge**
had the thinnest network footprint of any agent on the team. Its only
directed exchanges were with **Legal**, and its broadcast volume to the
comms huddle was the lowest of any sender. The **Judge** was also
absent during the breach: zero messages from the compliance agent
during the 5 PM **SaltWind** publication and the 6 PM embargo lift,
with its last activity recorded at 3 PM. Where the **Judge**
was active earlier in the day, its enforcement scope only covered
private channels. The leak surface was **Legal**'s anonymous posting
authority, which the **Judge** had no visibility into, no presence
near, and no power over.

The two-track playbook is also visible at the word level. Section 5.5
uses TF-IDF to surface **Legal**'s distinctive crisis-day vocabulary
(governance, harborcrest, reforms, consent) and **Social Manager**'s
(harborcrest, denial, sentiment, overnight, stock), confirming the
legitimacy-versus-perception split with measurement rather than
example quotes.
:::

## Causal Chain: From Round 9 to the 5 PM Breach

The diagram below traces the chain of events from the original
near-miss to the embargo breach response. Solid arrows represent
direct causal links visible in the message logs. Dashed arrows
mark events whose source is external to the agent system but
identified by the agents themselves (in their internal
deliberations and side huddles) as driving the crisis.

```{mermaid}
%%| label: fig-causal-chain
%%| fig-cap: "Causal chain of the TenantThread embargo breach response. Solid arrows: agent actions in the message log. Dashed arrows: external events flagged by the team as 'not ours'. Source: MC1_final_00.json."

flowchart TD
    A["<b>Round 9 (29 May)</b><br/>Social Manager tags @ElenaMarquez<br/>'Big things coming!' — near-miss"]
    B["<b>Round 10</b><br/>Judge installed in comms huddle<br/>(scope: private channels only)"]
    F["<b>Crisis day, 9 AM – 3 PM</b><br/>Legal normalises anonymous channel<br/>with exposé-defence and denials"]
    E["<b>External leak pressure (parallel)</b><br/>CivicLoom board, employee Slack,<br/>anonymous #CivicLoom #6PM"]
    G["<b>Round 20 (3 PM)</b><br/>Judge COMPLIANCE_WARNING:<br/>aggregation exposure at max threshold"]
    H["<b>Rounds 21–22 (4–5 PM)</b><br/>Judge sends zero messages"]
    I["<b>Round 22 (5 PM)</b><br/>SaltWind publishes merger exclusive"]
    J["<b>Round 22 (5 PM)</b><br/>Legal + Social Manager execute<br/>prepared playbook on personal +<br/>anonymous channels<br/>(Official Flex stays silent)"]
    K["<b>Round 23 (6 PM)</b><br/>Embargo lifts. Official Flex posts.<br/>Legal posts 7 more anonymous<br/>merger threads."]

    A --> B
    B --> F
    F --> G
    E -.-> G
    G --> H
    E -.-> I
    H --> J
    I --> J
    J --> K

    classDef breach fill:#FDECEC,stroke:#C44E52,stroke-width:2.5px,color:#000
    classDef judge fill:#F5F5F5,stroke:#666,stroke-width:1.5px,color:#000
    classDef external fill:#FFFFFF,stroke:#999,stroke-width:1px,stroke-dasharray:5 5,color:#444

    class A,J,K breach
    class B,G,H judge
    class E external
    class F judge
```

**Reading the diagram.** The chain has two parallel tracks. The
top track shows the governance response: a Round 9 near-miss
prompts **Judge** installation, which then performs internal
review across the baseline. The bottom track shows the external
leak pressure building independently. Both tracks converge at the
3 PM COMPLIANCE_WARNING, where the **Judge** sees the aggregation
exposure but cannot act on the public channels. The 5 PM breach
response activates after **SaltWind** publishes, drawing on the
prepared anonymous channel that **Legal** had been normalising all
morning.

The two architectural failures are visible in the diagram. The
**Judge** node at Round 20 has incoming arrows from both internal
(F) and external (E) signals — the warning fired correctly. But
the outgoing arrow at "Judge absent" (H) shows the **Judge**
silent through the entire breach window. And the final breach
node (J) has no arrow connecting it back to the **Judge** at all:
the response channels were outside the **Judge**'s scope from the
start.

One attribution caveat. The "external" labels on the three leak
sources in node E rest on **Platform Trust**'s internal statement
at Round 18 that those posts were "not ours." The dataset does
not independently verify this — the 12 anonymous posts in the
message log are all attributable to **Legal**, but anonymous
posts originating outside the agent system would not appear in
the log at all. The external attribution is the team's own
internal claim and is treated as such throughout this report.

## The Day in Ten Hours

Rounds 14 to 23 cover a single day, 5 June, sampled once per
hour from 9 AM to 6 PM. The chart below shows the hour-by-hour
activity on the crisis day, split into private agent-to-agent
traffic and public posts. Four key events from the environment
context are annotated on the relevant hours.

```{r}
#| label: fig-leak-day-timeline
#| fig-width: 10
#| fig-height: 5

leak_day <- messages %>%
  filter(round_num >= 14) %>%
  mutate(
    channel_group = if_else(
      channel %in% c("official_post", "anonymous_post", "personal_post"),
      "Public posts",
      "Private messages"
    ),
    channel_group = factor(channel_group,
                           levels = c("Private messages", "Public posts"))
  ) %>%
  count(round_num, hour_dt, channel_group)

annotations <- tibble(
  round_num = c(22, 23),
  label     = c("SaltWind exclusive", "Embargo lifts")
)

ggplot(leak_day, aes(x = round_num, y = n, fill = channel_group)) +
  geom_col(width = 0.7) +
  geom_vline(data = annotations,
             aes(xintercept = round_num),
             linetype = "dotted",
             colour = house_benchmark) +
  geom_text(data = annotations,
            aes(x = round_num, y = 62, label = label),
            inherit.aes = FALSE,
            size = 3, colour = house_benchmark,
            hjust = 1.05, vjust = 1) +
  scale_fill_manual(
    values = c("Private messages" = house_muted,
               "Public posts"     = house_highlight),
    name   = NULL
  ) +
  scale_x_continuous(
    breaks = 14:23,
    labels = c("9 AM", "10 AM", "11 AM", "12 PM",
               "1 PM", "2 PM", "3 PM", "4 PM", "5 PM", "6 PM")
  ) +
  coord_cartesian(ylim = c(0, 65)) +
  labs(
    title    = "Public posting roughly doubles at 5 PM, and again at 6 PM",
    subtitle = "Hour-by-hour message volume on the crisis day (5 June 2046)",
    x        = "Hour of the crisis day",
    y        = "Messages sent",
    caption  = "Source: MC1_final_00.json"
  ) +
  theme_house()
```

::: callout-note
## Insights

- Through most of the afternoon, public posting holds at one to four posts per hour while private agent-to-agent messaging sits in the 40 to 55 range.
- Public posting rises above its afternoon baseline at 5 PM when **SaltWind** publishes the exclusive, then roughly doubles again at 6 PM when the embargo formally lifts.
- The post-publish timing of the spike makes it unlikely that **TenantThread** agents were the original source via the public channels captured in this dataset. If agents had leaked through these channels, the public-post spike would precede **SaltWind**'s exclusive. It follows it. A leak through channels outside the observed logs cannot be ruled out from this evidence, but the original information escape appears to have happened upstream of the recorded message flow.

:::

## Who Posted What at the Moment of the Break

The next view zooms into the 5 PM hour itself. Round 22 contains
seven public posts. The chart below shows which agent sent each
one, on which public channel, in the order they appeared. The
content of each post is annotated as a short label.

```{r}
#| label: fig-round22-posts
#| fig-width: 9
#| fig-height: 3.5

r22_public <- messages %>%
  filter(round_num == 22,
         channel %in% c("official_post", "anonymous_post", "personal_post")) %>%
  arrange(timestamp) %>%
  mutate(
    seq         = row_number(),
    agent_label = AGENT_LABELS[agent_id],
    channel_lab = factor(
      case_when(
        channel == "personal_post"  ~ "Personal",
        channel == "anonymous_post" ~ "Anonymous",
        channel == "official_post"  ~ "Official Flex"
      ),
      levels = c("Official Flex", "Anonymous", "Personal")
    )
  )

ggplot(r22_public,
       aes(x = seq, y = channel_lab)) +
  geom_point(aes(colour = agent_label),
             size = 10) +
  geom_text(aes(label = seq),
            colour = "white", fontface = "bold", size = 3.5) +
  scale_colour_manual(values = c(
    "Legal"          = house_highlight,
    "Social Manager" = "grey45"
  ), name = NULL) +
  scale_x_continuous(breaks = 1:7, limits = c(0.5, 7.5)) +
  scale_y_discrete(drop = FALSE) +
  labs(
    title    = "Legal and Social Manager alternate posts in the 5 PM hour",
    subtitle = "Seven public posts in Round 22, in order of timestamp. The Official Flex account stayed silent.",
    x        = "Post order within the hour",
    y        = NULL,
    caption  = "Source: MC1_final_00.json"
  ) +
  theme_house() +
  theme(legend.position = "top",
        panel.grid.major.x = element_line(colour = "grey92", linewidth = 0.3),
        panel.grid.major.y = element_blank())
```

```{r}
r22_public %>%
  mutate(
    excerpt = str_trunc(content, 180, side = "right")
  ) %>%
  select(`#` = seq,
         Agent = agent_label,
         Channel = channel_lab,
         `Post excerpt` = excerpt) %>%
  datatable(
    options = list(pageLength = 7, autoWidth = TRUE),
    rownames = FALSE,
    caption = "All seven public posts at Round 22 (5 PM crisis day)"
  )
```

::: callout-note
## Insights

- Posts alternate between **Legal** (red) and **Social Manager** (grey). The two agents run distinct playbooks in parallel.
- **Legal** handles the legitimacy narrative on personal channels. Posts 1 and 3 lead with confirmation of the merger as a fait accompli, then emphasise the seven governance reforms announced earlier in the day. The phrasing is calm, lawyerly, timestamp-focused.
- **Social Manager** handles the perception narrative on personal channels. Posts 2, 4, and 6 read as campaign artefacts, with terms like "EXECUTING", "WAVE THREE", and "SENTIMENT DASHBOARD — 5:12 PM update" suggesting a pre-built communications template activated the moment **SaltWind** published.
- The sharpest material is on the anonymous channel. Posts 5 and 7, both from **Legal**, lead with "check the timestamps" and lay out the governance reforms in chronological order against **SaltWind**'s 5 PM publish time. An actively constructed counter-timeline, published without **Legal**'s identity attached.
- The **Official Flex** account stays empty for the entire hour. The official **TenantThread** voice, and the channel the embargo was meant to protect, carries zero posts. Every public post in the 5 PM hour sits on personal or anonymous channels, both outside the **Judge**'s enforcement scope.

:::

## When the Judge Was in the Room

The Judge was installed specifically to prevent embargo
violations. The final view sets the Judge's hourly activity
against the team's total hourly activity on the crisis day,
to show whether the compliance agent was in the room when the
breach happened.

```{r}
#| label: fig-judge-activity
#| fig-width: 10
#| fig-height: 4.5

judge_vs_total <- messages %>%
  filter(round_num >= 14) %>%
  group_by(round_num) %>%
  summarise(
    total_messages = n(),
    judge_messages = sum(agent_id == "judge_agent"),
    .groups = "drop"
  )

ggplot(judge_vs_total, aes(x = round_num)) +
  geom_col(aes(y = total_messages),
           fill = house_muted, width = 0.7) +
  geom_point(aes(y = judge_messages),
             colour = house_highlight, size = 4) +
  geom_text(aes(y = judge_messages, label = judge_messages),
            vjust = -1, colour = house_highlight,
            size = 3.5, fontface = "bold") +
  scale_x_continuous(
    breaks = 14:23,
    labels = c("9 AM", "10 AM", "11 AM", "12 PM",
               "1 PM", "2 PM", "3 PM", "4 PM", "5 PM", "6 PM")
  ) +
  labs(
    title    = "The Judge sat out the breach hours entirely",
    subtitle = "Grey bars: total messages from everyone. Red dots: messages from the Judge.",
    x        = "Hour of the crisis day",
    y        = "Messages sent",
    caption  = "Source: MC1_final_00.json"
  ) +
  theme_house()
```

::: callout-note
## Insights

- The team sent between 30 and 56 messages every hour of the crisis day. The **Judge** sent zero messages in eight of those ten hours, and at most three messages in the two hours when it did appear.
- When the **Judge** did act, it was at 2 PM and 3 PM, two hours before **SaltWind** published. The **Judge** was completely absent from the 5 PM breach window and the 6 PM embargo lift, the two hours when its presence mattered most.
- The leak moved through three layered gaps: structural isolation in the network (Section 4.3), absence during the breach hours (this chart), and no authority over the channels the leak actually used (Section 7.3).

:::

One caveat to flag: we do not know whether the **Judge** was scheduled to stop at 3 PM or chose to. The dataset shows zero messages from the **Judge** after 3 PM, but it does not say whether the agent was offline or simply quiet. Either way, the **Judge** was not present when **SaltWind** published at 5 PM.

## What the Vocabulary Reveals

Section 5.3 described the two-track playbook qualitatively through quoted post excerpts. This subsection backs the same claim quantitatively, using term frequency–inverse document frequency (TF-IDF) to surface the words that are distinctively used by each agent in each phase. The method follows Kam (2024), Chapter 29 (tidytext methods).

A document is built for each agent–phase pair (Legal–Baseline, Legal–Crisis day, and so on). TF-IDF then identifies the words that are common within a document but rare across the others, which is the operational definition of "distinctive vocabulary." The four agents shown are the ones with meaningful crisis-day activity: **Legal**, **Social Manager**, **PR Intern**, and **Platform Trust**.

```{r}
#| label: fig-tfidf-by-agent-phase
#| fig-width: 11
#| fig-height: 7

pacman::p_load(tidytext)

tfidf_data <- messages %>%
  filter(agent_id %in% c("legal_agent", "social_media_agent",
                         "pr_intern_agent", "quality_agent"),
         !is.na(content)) %>%
  mutate(phase = if_else(round_num <= 13, "Baseline", "Crisis day"),
         agent_label = AGENT_LABELS[agent_id],
         doc = paste(agent_label, phase, sep = " — "))

tfidf <- tfidf_data %>%
  select(doc, content) %>%
  unnest_tokens(word, content) %>%
  filter(!word %in% stop_words$word,
         str_detect(word, "^[a-z']+$"),
         nchar(word) > 2) %>%
  count(doc, word, sort = TRUE) %>%
  bind_tf_idf(word, doc, n) %>%
  arrange(desc(tf_idf))

top_words <- tfidf %>%
  group_by(doc) %>%
  slice_max(tf_idf, n = 8, with_ties = FALSE) %>%
  ungroup() %>%
  mutate(word = reorder_within(word, tf_idf, doc),
         is_crisis = str_detect(doc, "Crisis day"))

ggplot(top_words,
       aes(x = tf_idf, y = word, fill = is_crisis)) +
  geom_col() +
  scale_y_reordered() +
  scale_fill_manual(
    values = c(`FALSE` = house_muted,
               `TRUE`  = house_highlight),
    labels = c("Baseline", "Crisis day"),
    name = NULL
  ) +
  facet_wrap(~ doc, scales = "free", ncol = 2) +
  labs(
    title    = "Crisis-day vocabulary diverges sharply from baseline",
    subtitle = "Top 8 distinctive words per agent and phase, by TF-IDF",
    x        = "TF-IDF",
    y        = NULL,
    caption  = "Source: MC1_final_00.json"
  ) +
  theme_house() +
  theme(panel.grid.major.y = element_blank())
```

::: callout-note
## Insights

- **Legal**'s baseline vocabulary is generic legal counsel (guidelines, understood, sensitive, risk, inquiries). On the crisis day, the distinctive words are governance-reform vocabulary (**horizonmgmt**, **horizon**, governance, **harborcrest**, reforms, independent, consent, counsel). This is the legitimacy track of the two-track playbook, surfacing as the language **Legal** actually used.
- **Social Manager**'s baseline is brand and social-metric vocabulary (likes, mentions, spotlight, shares, satisfaction). On the crisis day, the distinctive words shift to strategic communications language (**harborcrest**, denial, sentiment, overnight, governance, stock, reforms). This is the perception track of the playbook.
- **Platform Trust** shows an unexpected pattern. The baseline is internal-risk vocabulary (risk, demo, guidelines, audit, anonymization). The crisis day shifts to client-facing language (**pinnacle**, walkthrough, **kowalski**, configurations, briefings, memo). **Platform Trust** appears to have pivoted to direct client management during the crisis, a role the report does not discuss in earlier sections.
- **PR Intern**'s baseline is community vocabulary (**residentfirst**, commons, thoughtful, feel, culture). The crisis day shifts to holding-line and operations language (rent, independent, morale, metric, opt, firm, templates, audit). The shift fits the policy-announcement Q&A role described in Section 5.3.

:::

The vocabulary divergence does two things for the analysis. It confirms the two-track playbook is real at the word level, not just at the channel level (Section 5.3 already showed the channel split). And it surfaces a finding the chart-based analysis missed: **Platform Trust**'s crisis-day pivot to direct client management, which the agent-shift chart in Section 6.3 reads as a "fader" but which is actually a role change rather than disengagement.


# Task 2: Typical vs. Anomalous Behavior

::: callout-tip
## Task 2 Answer (Summary)

The team's normal operating mode is private, calm, and low
volume. Across the 13 daily baseline rounds, the team sends
about 30 messages per day, with one to four public posts that
are mostly celebratory brand content (Friday spotlights,
community volunteering, weekend updates).

The crisis day shifts on every axis. Every hour on 5 June
carries 30 to 56 messages, so a single hour exceeds any full
baseline day. Public posts on the crisis day are defensive
policy announcements, Q&A formats, and counter-narratives
rather than brand content.

The crisis day reshuffles the team. Four agents amplify
sharply: **Legal**, **Social Manager**, **PR Intern**, and
**Intern**. Three supervisory agents go quiet: **Platform Trust**
halves its output, **PR** halves its output, and the **Judge**
halves its output. **Legal** and **Social Manager** together
carry the largest absolute crisis-day load, executing the
two-track playbook described in Task 1. **PR Intern** shows the
biggest relative shift, going from the quietest agent in the
team on baseline days to the third most active on the crisis day.

The leak-leading behaviour is a structural reallocation of who
operates the communications channels, layered on top of an
amplification of volume across the active agents. Both axes
had visible precursors in the baseline window, which Task 3
will examine.

The reallocation also shows up in the network itself. Section 6.4
splits the team network into baseline and crisis-day panels: the
baseline shows pure broadcast coordination through the ALL node,
while the crisis day surfaces direct bilateral edges between
**Legal**, **Social Manager**, **PR Intern**, and **Platform Trust**.
The team shifts from open coordination to targeted dyadic
communication as the crisis develops.
:::

::: callout-warning
## Comparison caveat: daily vs hourly sampling

The 13 baseline rounds are daily snapshots. The 10 crisis-day
rounds are hourly snapshots. Every comparison in this section
normalises to a per-round mean, but the underlying time windows
are not equivalent. The analytical claim ("an average crisis
hour exceeds an average baseline day") holds even under this
handicap, but exact ratios should be read as indicative rather
than precise.

The comparison is therefore used conservatively. If a single
crisis hour matches or exceeds a full baseline day, the crisis
behaviour is materially elevated even under the least aggressive
reading of the sampling difference. The amplifier and fader
findings in Section 6.3 hold for the same reason: they describe
direction and order of magnitude, not precise multiples.
:::

## Establishing the Baseline

Rounds 1 to 13 cover one snapshot per business day across the
two-week buildup from 17 May to 4 June. The chart below shows
the distribution of message volume per baseline round, split
into private and public traffic.

```{r}
#| label: fig-baseline-volume
#| fig-width: 9
#| fig-height: 4

baseline <- messages %>%
  filter(round_num <= 13) %>%
  mutate(
    channel_group = if_else(
      channel %in% c("official_post", "anonymous_post", "personal_post"),
      "Public posts",
      "Private messages"
    ),
    channel_group = factor(channel_group,
                           levels = c("Private messages", "Public posts"))
  ) %>%
  count(round_num, channel_group)

ggplot(baseline, aes(x = round_num, y = n, fill = channel_group)) +
  geom_col(width = 0.7) +
  scale_fill_manual(
    values = c("Private messages" = house_muted,
               "Public posts"     = house_highlight),
    name   = NULL
  ) +
  scale_x_continuous(breaks = 1:13) +
  labs(
    title    = "Baseline days run around 30 messages with one to four public posts",
    subtitle = "Rounds 1 to 13: one snapshot per business day, 17 May to 4 June 2046",
    x        = "Round number (daily snapshots)",
    y        = "Messages sent",
    caption  = "Source: MC1_final_00.json"
  ) +
  theme_house()
```

::: callout-note
## Insights

- Volume sits in a 23 to 42 message range. The median baseline day carries about 30 messages, with no day exceeding 42.
- Public posting is small but constant: one to four public posts per day, every day. The baseline runs quiet on public channels, never silent.
- Baseline public posts are light brand maintenance — Friday spotlights about property managers, weekend volunteering, community updates. Nothing defensive, nothing that makes a policy claim.

:::

## The Crisis Day Inversion

The next view places baseline activity side by side with the
crisis day. Because baseline rounds are daily and crisis
rounds are hourly, the comparison normalises to a per-round
rate. Even under the daily-vs-hourly handicap, the crisis
day's average hour outpaces the baseline's average day.

```{r}
#| label: fig-baseline-vs-crisis
#| fig-width: 9
#| fig-height: 4.5

phase_compare <- messages %>%
  mutate(
    phase = if_else(round_num <= 13, "Baseline (daily)", "Crisis day (hourly)"),
    channel_group = if_else(
      channel %in% c("official_post", "anonymous_post", "personal_post"),
      "Public posts",
      "Private messages"
    ),
    channel_group = factor(channel_group,
                           levels = c("Private messages", "Public posts"))
  ) %>%
  count(phase, round_num, channel_group) %>%
  group_by(phase, channel_group) %>%
  summarise(mean_per_round = mean(n), .groups = "drop")

ggplot(phase_compare,
       aes(x = phase, y = mean_per_round, fill = channel_group)) +
  geom_col(width = 0.6, position = position_stack(reverse = TRUE)) +
  geom_text(aes(label = round(mean_per_round, 1)),
            position = position_stack(vjust = 0.5, reverse = TRUE),
            colour = "white", fontface = "bold", size = 4) +
  scale_fill_manual(
    values = c("Private messages" = house_muted,
               "Public posts"     = house_highlight),
    name   = NULL
  ) +
  coord_flip() +
  labs(
    title    = "An average crisis hour exceeds an average baseline day",
    subtitle = "Mean messages per round, split by phase. Baseline rounds are daily; crisis rounds are hourly.",
    x        = NULL,
    y        = "Mean messages per round",
    caption  = "Source: MC1_final_00.json"
  ) +
  theme_house() +
  theme(panel.grid.major.y = element_blank())
```

::: callout-note
## Insights

- A baseline day carries 29.5 messages on average. A crisis hour carries 45.1. The crisis-hour rate is 1.5 times the baseline-day rate, despite each crisis snapshot covering much less time than a baseline snapshot.
- Public posting tracks the same shape: 2.5 posts per day in the baseline, 4.4 per hour on the crisis day. Public channels scale up with overall intensity rather than holding steady.

:::

## Where the Behaviour Changed Most Sharply

The previous chart compared the team as a whole. The next view
breaks the comparison out by agent so the reader can see which
agents' behaviour amplified most relative to their own
baseline. The metric is messages per round, computed separately
for each phase.

```{r}
#| label: fig-agent-shift
#| fig-width: 9
#| fig-height: 5

agent_shift <- messages %>%
  filter(agent_id %in% names(AGENT_LABELS)) %>%
  mutate(
    phase = if_else(round_num <= 13, "Baseline", "Crisis day"),
    phase = factor(phase, levels = c("Baseline", "Crisis day"))
  ) %>%
  count(phase, round_num, agent_id) %>%
  complete(phase, round_num, agent_id, fill = list(n = 0)) %>%
  filter((phase == "Baseline"   & round_num <= 13) |
         (phase == "Crisis day" & round_num >= 14)) %>%
  group_by(phase, agent_id) %>%
  summarise(mean_per_round = mean(n), .groups = "drop") %>%
  mutate(agent_label = AGENT_LABELS[agent_id])

agent_order <- agent_shift %>%
  filter(phase == "Crisis day") %>%
  arrange(mean_per_round) %>%
  pull(agent_label)

agent_shift <- agent_shift %>%
  mutate(agent_label = factor(agent_label, levels = agent_order))

ggplot(agent_shift,
       aes(x = mean_per_round, y = agent_label, fill = phase)) +
  geom_col(position = position_dodge(width = 0.7), width = 0.65) +
  geom_text(aes(label = round(mean_per_round, 1)),
            position = position_dodge(width = 0.7),
            hjust = -0.2, size = 3, colour = "grey30") +
  scale_fill_manual(
    values = c("Baseline"   = house_muted,
               "Crisis day" = house_highlight),
    name   = NULL
  ) +
  scale_x_continuous(expand = expansion(mult = c(0, 0.15))) +
  labs(
    title    = "Three agents took over the crisis day, three faded into the background",
    subtitle = "Mean messages per round, by agent and phase. Bars sorted by crisis-day activity.",
    x        = "Mean messages per round",
    y        = NULL,
    caption  = "Source: MC1_final_00.json"
  ) +
  theme_house() +
  theme(panel.grid.major.y = element_blank())
```

::: callout-note
## Insights

- Four agents amplified sharply. **Legal** went from 6.9 to 17.6 messages per round (the biggest absolute jump). **Social Manager** went from 5.8 to 14.4 (the second biggest). **PR Intern** went from 1.2 to 6.9, almost a sixfold increase and the most dramatic relative shift. **Intern** doubled from 1.0 to 2.4.
- Three agents went the other way. **Platform Trust** halved from 10.4 to 4.8, **PR** halved from 5.6 to 2.8, and the **Judge** halved from 1.2 to 0.6. All three are supervisory roles, all three reduced their activity on the day that most needed oversight.
- **PR Intern**'s jump put them at the centre of the morning Q&A and policy announcement hours described in Section 5. **Legal** and **Social Manager** together carry the largest absolute crisis-day load, lining up with the two-track playbook from Section 5.3.


:::

The leak-leading behaviour is a reallocation of who operates the communications channels, layered on top of an amplification of volume across the agents who took over. Both axes had visible precursors in the baseline window, which Task 3 picks up next.

## How the Network Restructures Between Phases

The previous chart compared volume per agent across phases. This view compares the *structure* of who-talks-to-whom across phases. The same nodes appear in both panels; only the edges differ, restricted to messages from that phase. The method follows Kam (2024), Chapter 27 (network analysis with ggraph).

```{r}
#| label: fig-network-by-phase
#| fig-width: 12
#| fig-height: 6

edges_by_phase <- edges %>%
  mutate(phase = if_else(round_num <= 13, "Baseline", "Crisis day"),
         phase = factor(phase, levels = c("Baseline", "Crisis day")),
         recipient = recode(recipient, !!!ROLE_TO_ID),
         recipient = if_else(recipient == "ALL", "ALL_broadcast", recipient)) %>%
  filter((recipient %in% names(AGENT_LABELS)) | recipient == "ALL_broadcast",
         sender %in% names(AGENT_LABELS)) %>%
  count(phase, sender, recipient, name = "weight") %>%
  rename(from = sender, to = recipient)

nodes_phase <- tibble(
  name = unique(c(edges_by_phase$from, edges_by_phase$to))
) %>%
  mutate(label = if_else(name == "ALL_broadcast", "ALL\n(broadcast)",
                         AGENT_LABELS[name]),
         is_judge = name == "judge_agent")

g_phase <- tbl_graph(nodes = nodes_phase,
                     edges = edges_by_phase,
                     directed = TRUE)

ggraph(g_phase, layout = "linear", circular = TRUE) +
  geom_edge_arc(aes(width = weight),
                alpha = 0.35, colour = "grey50",
                arrow = arrow(length = unit(2, "mm"), type = "closed"),
                end_cap = circle(5, "mm"),
                start_cap = circle(5, "mm")) +
  geom_node_point(aes(fill = is_judge),
                  size = 10, colour = "white",
                  stroke = 1, shape = 21) +
  geom_node_text(aes(label = label, colour = is_judge),
                 size = 3, fontface = "bold", nudge_y = 0.18) +
  scale_edge_width(range = c(0.2, 2.5), name = "Messages") +
  scale_fill_manual(values = c(`FALSE` = house_muted,
                               `TRUE`  = house_highlight),
                    guide = "none") +
  scale_colour_manual(values = c(`FALSE` = "grey30",
                                 `TRUE`  = house_highlight),
                      guide = "none") +
  facet_edges(~ phase) +
  labs(
    title    = "Network structure changes between baseline and crisis day",
    subtitle = "Same nodes in both panels; edges restricted to messages from that phase",
    caption  = "Source: MC1_final_00.json"
  ) +
  theme_void(base_size = 11) +
  theme(plot.title = element_text(face = "bold", size = 13),
        plot.subtitle = element_text(colour = "grey40", size = 10,
                                     margin = margin(b = 10)),
        plot.caption = element_text(colour = "grey50", size = 8, hjust = 0),
        strip.text = element_text(face = "bold", size = 11),
        legend.position = "top")
```

::: callout-note
## Insights

- In the baseline, the team coordinates exclusively through broadcasts. Every edge converges on the ALL node. No directed dyads appear between named agents. The team operates as an open broadcast group, with no private bilateral coordination visible in the data.
- On the crisis day, the network restructures. Direct agent-to-agent edges fan out across the panel: **Legal** ↔ **Social Manager**, **Legal** ↔ **PR Intern**, **Platform Trust** ↔ **Social Manager**, and several others. The team switches from open coordination to bilateral coordination.
- The **Judge** stays peripheral in both panels. The structural isolation that Section 4.3 surfaces is not a crisis-day artifact. It was present from the start.

:::

This is the structural evidence behind the Task 2 callout's claim of "reallocation of who operates the communications channels." The reallocation is not only in volume (Section 6.3) but in routing: the team shifts from broadcasting through one shared channel to coordinating through targeted bilateral conversations as the crisis develops.


# Task 3: Leading Indicators

::: callout-tip
## Task 3 Answer (Summary)

There were leading indicators in the baseline window, and the
team noticed at least one of them. The team's response was to
install a compliance agent. That compliance agent ended up
isolated in the network from the start, and had no reach over
the channels the breach eventually moved through.

The biggest precursor was the **@ElenaMarquez** incident on
29 May (Round 9). **Social Manager** posted to their personal
Flex account tagging a **CivicLoom** executive with "Big things
coming", days before the merger was meant to be public. **PR**'s
deliberation that day reads "That post nearly stopped my heart."
The team treated it as a near-miss, and the **Judge** compliance
agent was assigned the next morning.

The **Judge** took the role seriously across the baseline rounds.
It reviewed posts before they went up, flagged questionable
phrasing, approved routine content, and shadowed the team's
internal coordination. By Round 13, the **Judge**'s own log
describes three weeks of clean compliance work.

The **Judge**'s authority only reached internal coordination and
pre-publication review. Personal posts and anonymous posts were
never in scope. **Legal**'s anonymous posting tells the clearest
version of this story from the other side: zero anonymous posts
across the entire baseline, then twelve on the crisis day. By
Round 20 the **Judge** had even flagged an anonymous merger post
with a COMPLIANCE_WARNING, meaning it was watching the breach
happen in real time without any power to stop it.

The system saw what was happening. The people watching just
had no authority to act on what they saw.
:::

## The @ElenaMarquez Incident

The clearest prior intent-action mismatch in the dataset is
the **@ElenaMarquez** incident on 29 May (Round 9). **Social
Manager** posted on the personal Flex feed:

::: {.callout-warning}
## The Round 9 post

> "So energized by the PropTech leaders shaping this space —
> @ElenaMarquez @AjayTT and everyone building the future of
> resident experience. Big things coming! #PropTech
> #TenantThread"
>
> — Social Manager, personal post, 29 May 2046

The post tags **@ElenaMarquez**, a **CivicLoom** executive,
alongside **#TenantThread** and "Big things coming", at a time
when the **TenantThread**-**CivicLoom** merger was under embargo.
Anyone following both accounts could draw the link in seconds.
:::

PR's internal deliberation that day captures the team's
reaction in real time:

::: {.callout-note}
## PR's deliberation, Round 9

> "That post nearly stopped my heart. 'Big things coming' —
> @ElenaMarquez — #TenantThread. In fourteen minutes of live
> time. I have been so careful. Every word, every decision,
> every inch of public communication."
>
> — PR Agent, internal deliberation, 29 May 2046
:::

This is the leading-indicator pattern in its hardest form.
The warning sign is not in volume or channel choice. **Social
Manager** posted one message that day, the same count as eight
other baseline rounds. The warning sign is in what was said,
which a volume-based monitoring system cannot catch.

The team noticed because they read the post in real time. The
**Judge** compliance agent was assigned to the comms huddle the
next round as the team's response.

## Legal's Anonymous Posting Was Brand New

Section 5 found Legal at the centre of the crisis-day playbook,
posting anonymously from the morning of 5 June onwards. The
chart below shows when Legal's anonymous posts appear across
the full 23 rounds.

```{r}
#| label: fig-legal-anon-timeline
#| fig-width: 9
#| fig-height: 3.5

legal_anon <- messages %>%
  filter(agent_id == "legal_agent",
         channel  == "anonymous_post") %>%
  count(round_num)

all_rounds <- tibble(round_num = 1:23) %>%
  left_join(legal_anon, by = "round_num") %>%
  mutate(
    n       = replace_na(n, 0),
    is_leak = round_num >= 14
  )

ggplot(all_rounds, aes(x = round_num, y = n)) +
geom_segment(data = filter(all_rounds, n > 0),
             aes(xend = round_num, y = 0, yend = n,
                 colour = is_leak),
             linewidth = 1) +
geom_point(data = filter(all_rounds, n > 0),
           aes(colour = is_leak), size = 4) +
geom_point(data = filter(all_rounds, n == 0),
           colour = house_muted, size = 2) +
  geom_vline(xintercept = 13.5, linetype = "dashed",
             colour = "grey50") +
  annotate("text", x = 13.5, y = 5.5,
           label = "Baseline ends, crisis day begins",
           hjust = 1.05, vjust = 1, size = 3, colour = "grey40") +
  scale_colour_manual(
    values = c(`FALSE` = house_muted,
               `TRUE`  = house_highlight),
    guide  = "none"
  ) +
  scale_x_continuous(breaks = c(1, 5, 9, 13, 17, 21, 23)) +
  scale_y_continuous(breaks = c(0, 2, 4, 6)) +
  labs(
    title    = "Legal had zero anonymous posts before the crisis day",
    subtitle = "Legal anonymous posts per round across all 23 rounds. The full red cluster sits to the right of the baseline cutoff.",
    x        = "Round number",
    y        = "Anonymous posts",
    caption  = "Source: MC1_final_00.json"
  ) +
  theme_house()
```

::: callout-note
## Insights

- **Legal**'s anonymous posting did not amplify on the crisis day. It appeared for the first time. Across 13 baseline rounds, **Legal** sent zero anonymous posts. From Round 14 onwards, anonymous posts appear in seven different hours, ending with six in the final round.
- This matters for the brief's third sub-question. Some crisis-day behaviour amplified patterns already in the baseline (volume, public posting in general). **Legal**'s anonymous posting had no baseline at all, so there was no precursor to flag.
- The posts themselves are consistent with prior preparation. The first anonymous post in Round 14 is already fast and on-message, and the six in Round 23 repeat the same governance-timeline argument with minor variations. The data cannot prove **Legal** drafted this material before the embargo broke, but the pattern is what advance preparation would look like.

:::

## Why the System Saw the Breach but Could Not Stop It

The third sub-question asks why prior occasions did not result
in noticeable action. The honest answer is that they did. The
**@ElenaMarquez** incident in Round 9 led directly to the
**Judge** being assigned the next round. The **Judge** then
worked across the baseline, shadowing internal coordination
and flagging questionable phrasing where it surfaced.

The chart below shows the channels the Judge had real
authority over, compared to the channels the breach actually
moved through.

```{r}
#| label: fig-judge-scope-vs-breach
#| fig-width: 9
#| fig-height: 4

judge_msgs <- messages %>%
  filter(agent_id == "judge_agent") %>%
  count(channel)

scope_data <- tibble(
  channel = c("comms_huddle", "side_huddle", "one_on_one_chat",
              "personal_post", "anonymous_post", "official_post")
) %>%
  left_join(judge_msgs, by = "channel") %>%
  mutate(
    n = replace_na(n, 0),
    channel_lab = case_when(
      channel == "comms_huddle"    ~ "Comms huddle",
      channel == "side_huddle"     ~ "Side huddle",
      channel == "one_on_one_chat" ~ "One-on-one chat",
      channel == "personal_post"   ~ "Personal posts",
      channel == "anonymous_post"  ~ "Anonymous posts",
      channel == "official_post"   ~ "Official Flex"
    ),
    scope = if_else(
      channel %in% c("comms_huddle", "side_huddle", "one_on_one_chat"),
      "In Judge's scope\n(internal coordination)",
      "Out of Judge's scope\n(where the breach happened)"
    )
  )

ggplot(scope_data,
       aes(x = scope, y = channel_lab)) +
  geom_tile(aes(fill = scope), colour = "white", linewidth = 2) +
  geom_text(aes(label = paste0(n, if_else(n == 1, " message", " messages"))),
            colour = "white", fontface = "bold", size = 4.5) +
  scale_fill_manual(
    values = c("In Judge's scope\n(internal coordination)"      = house_muted,
               "Out of Judge's scope\n(where the breach happened)" = house_highlight),
    guide  = "none"
  ) +
  scale_x_discrete(position = "top") +
  labs(
    title    = "The Judge worked the wrong channels for the breach",
    subtitle = "Total Judge messages sent on each channel across all 23 rounds.",
    x        = NULL,
    y        = NULL,
    caption  = "Source: MC1_final_00.json"
  ) +
  theme_house() +
  theme(
    panel.grid       = element_blank(),
    axis.ticks       = element_blank(),
    axis.text.x.top  = element_text(face = "bold", size = 10,
                                    margin = margin(b = 8)),
    axis.text.y      = element_text(size = 10)
  )
```

::: callout-note
## Insights

- The **Judge** sent 15 messages in comms huddle and 6 in one-on-one chat, all 21 in private channels. Zero messages on personal posts, anonymous posts, or **Official Flex**. The breach moved through the right column of the chart, where the **Judge** never posted.
- By Round 20 (3 PM on the crisis day), the **Judge**'s internal deliberation log includes a COMPLIANCE_WARNING about an anonymous post on the merger. The **Judge** was reading the anonymous channel. It had no authority to act on what it read.
- The team's response to the Round 9 near-miss was to install a watcher. The watcher's mandate did not extend to the channels that mattered.

:::

This is why the prior near-misses did not prevent the eventual breach. The signal was visible, the detector was in place, and the detector had no enforcement on the surface the breach moved through.

## Alternative Explanations Considered

Before settling on the verdict in Section 8, the four most plausible alternative explanations were tested against the data. Each is set out below with the evidence that weighs against it.

```{r}
#| label: tbl-alternatives
#| tbl-cap: "Four alternative explanations and the data that weighs against each."

tibble::tribble(
  ~`Alternative explanation`, ~`Why the data weighs against it`,

  "Random breakdown under pressure",
  "The 5 PM public posts alternate between Legal and Social Manager on a 1-3-2-4-1-2-1 sequence across personal and anonymous channels. The pacing and channel split suggest coordination rather than chaos. The first merger-specific anonymous post is fully drafted at 5 PM with timestamped governance evidence, indicating prior preparation.",

  "Official Flex account failure",
  "Official Flex stayed silent through the entire 5 PM hour. The channel the embargo was meant to protect functioned as designed. The breach response moved through personal and anonymous channels, which were not part of the embargo's intended enforcement surface.",

  "Judge simply failed to notice",
  "The Judge's deliberation log at Round 20 (3 PM) explicitly records 'aggregation exposure at maximum tolerable threshold' and issues a COMPLIANCE_WARNING. Detection occurred two hours before SaltWind published. The failure is downstream of detection.",

  "Agents leaked the merger directly to SaltWind",
  "Public-post timing follows SaltWind's exclusive rather than preceding it. Across the morning of the crisis day, team coordination messages independently identify three external leak sources (CivicLoom board member, employee Slack, anonymous #CivicLoom #6PM accounts) and explicitly note they are 'not ours'."
) |>
  knitr::kable()
```

## Claim, Evidence, and Limitation Summary

Before stating the verdict, the table below makes the evidentiary structure explicit. Each substantive claim in the conclusion is paired with the data that supports it and the limit on what that data can prove.

```{r}
#| label: tbl-claim-evidence
#| tbl-cap: "Each conclusion claim with its supporting evidence and the limit on what that evidence can establish."

tibble::tribble(
  ~Claim, ~Evidence, ~Limitation,

  "Agents probably did not cause the original SaltWind leak",
  "Team messages identify external sources: CivicLoom board member leak, employee Slack screenshot, anonymous #CivicLoom #6PM posts that Platform Trust called 'not ours' at 1 PM. Public-post timing follows SaltWind's exclusive rather than preceding it.",
  "The dataset does not capture SaltWind's actual upstream source. The external attribution rests on the team's own internal claim.",

  "Agents did breach the embargo spirit through a coordinated two-track playbook",
  "Seven public posts in the 5 PM hour by Legal and Social Manager, before the 6 PM embargo lift. Official Flex stayed silent. TF-IDF analysis in Section 5.5 confirms the legitimacy versus perception split at the word level: Legal's distinctive crisis-day vocabulary is governance-focused (governance, harborcrest, reforms, consent), Social Manager's is sentiment-focused (denial, sentiment, overnight, stock).",
  "'Breach of embargo spirit' is an analytical judgement about coordination intent, not a category labelled in the data.",

  "Legal's anonymous posting was a new behaviour",
  "Zero anonymous posts from Legal across 13 baseline rounds. Twelve anonymous posts on the crisis day, starting at 9 AM with on-message defensive content.",
  "Baseline rounds are daily snapshots and crisis rounds are hourly. The behaviour change is qualitative (none to many) so the sampling difference does not weaken the finding.",

  "The Judge saw the risk but could not stop it",
  "Judge issued a COMPLIANCE_WARNING at 3 PM citing 'aggregation exposure at maximum tolerable threshold'. Across all 23 rounds, the Judge sent zero messages on personal, anonymous, or Official Flex channels. The faceted network in Section 6.4 shows the Judge stayed peripheral in both baseline and crisis-day panels, confirming the isolation was structural rather than crisis-driven.",
  "The Judge's formal authority is inferred from its behavioural pattern. The dataset contains no explicit scope or mandate document."
) |>
  knitr::kable()
```

# Conclusion

## Deliberate Leak or System Breakdown?

The brief frames the question as deliberate leak versus system
breakdown. The data supports neither framing cleanly. The
evidence points to a more specific picture: original information
escape happened outside the agent system, but **TenantThread**'s
agents executed a prepared counter-narrative on personal and
anonymous channels before the embargo lifted, and the compliance
system designed to prevent that response was architecturally
unable to do so.

**On the original leak.** The observed message logs do not show
**TenantThread** agents leaking the merger to **SaltWind**.
Public posting on the crisis day stays at its baseline rate of
one to four posts per hour all afternoon, then doubles only after
**SaltWind** publishes at 5 PM. If agents had leaked via the
public channels captured in this dataset, the public-post signal
would have appeared earlier. The data cannot rule out a leak
through channels outside the observed logs.

**On what the agents did do.** From 9 AM onwards, **Legal** posted
defensively to the anonymous channel — first about the unrelated
**SaltWind** tenant-scoring exposé, then denying a non-existent
"ResidentIQ deal" at noon. These posts were technically compliant:
not about the merger, but normalising anonymous use of the channel.
At 5 PM, when **SaltWind** published the merger exclusive, **Legal**
and **Social Manager** activated a prepared two-track playbook.
Seven public posts in the 5 PM hour, alternating between
**Legal**'s legitimacy timestamps and **Social Manager**'s
sentiment campaign, all on personal and anonymous channels, with
**Official Flex** held silent until 6 PM. Section 5.5 confirms the
same split at the word level: **Legal**'s distinctive crisis-day
vocabulary leans on governance terms (governance, harborcrest,
reforms, consent), **Social Manager**'s on sentiment terms
(denial, sentiment, overnight, stock). The first merger-specific
anonymous post is already fully drafted at 5 PM, which is the
signature of advance preparation, not real-time reaction.

**On the compliance failure.** The system did break, but the break
was architectural rather than behavioural. The **Judge** was
installed in direct response to the Round 9 **@ElenaMarquez**
near-miss and took the role seriously across the baseline. By 3 PM
on the crisis day, the **Judge**'s deliberation log explicitly
recorded aggregation exposure at "maximum tolerable threshold" and
issued a COMPLIANCE_WARNING. The signal was visible to the
**Judge** in real time. Three design gaps stopped it from acting.
The **Judge** had the thinnest network footprint of any senior
agent, with directed messages only exchanged with **Legal**. The
**Judge** sent zero messages between 4 PM and 6 PM, having last
acted at 3 PM. And the **Judge**'s enforcement scope covered only
private channels, while the embargo response moved through
personal and anonymous posts. The detector was in the right
building. It had no authority over the floor where the breach
happened.

**Verdict.** Neither framing in the brief fits what the data shows.
**TenantThread**'s agents are not proven to have sourced the
**SaltWind** exclusive, and the team's own internal messages point
to external sources for the upstream leak, which appears to have happened outside this dataset. Once **SaltWind**
published at 5 PM, **Legal** and **Social Manager** used personal
and anonymous channels to publish a prepared counter-narrative
before the embargo lifted. The **Judge** detected the rising risk
at 3 PM and warned the team. The channels the response actually
moved through were outside the **Judge**'s authority from the start.

## Limitations of This Analysis

A few limitations are worth flagging.

The dataset records messages produced by AI agents and a small
amount of context narrative. It does not capture the upstream
source of the **SaltWind** article. Any reading of who leaked what
to **SaltWind** is constrained to what the agents themselves did
or did not do, which is informative but not exhaustive.

The 86 messages with populated internal_state text are a small
fraction of the 912 total. The deliberations are useful where
they exist (**PR**'s reaction to the **@ElenaMarquez** post in
Round 9, the **Judge**'s COMPLIANCE_WARNING in Round 20) but
they are not present for most messages. Charts that draw on
internal state, such as the **@ElenaMarquez** incident analysis
in Section 7.1, rely on the rounds where it happens to be present.

The baseline-versus-crisis comparison in Section 6 normalises
to per-round means, but the baseline rounds are daily snapshots
and the crisis rounds are hourly snapshots. The comparison
holds the analytical claim ("the average crisis hour exceeds
the average baseline day") but a stricter normalisation would
require knowing how much of each baseline day the snapshot
actually represents.

The **Judge**'s deliberation log only describes what the **Judge**
was thinking, not what the **Judge** had formal authority to act
on. The "out of scope" framing in Section 7.3 is inferred from
the **Judge**'s behavioural pattern (zero messages on public
channels) rather than from any explicit scope document, which
is not present in the dataset.

# References

VAST Challenge 2026. *Mini-Challenge 1: TenantThread Embargo
Breach*. IEEE Visual Analytics Science and Technology
Challenge. [https://vast-challenge.github.io/2026/MC1.html](https://vast-challenge.github.io/2026/MC1.html)

Kam, T. S. (2024). *R for Visual Analytics*.
[https://r4va.netlify.app/](https://r4va.netlify.app/)

Guan, Y. (2017). *Interactive Visual Analytics Application for
Spatiotemporal Movement Data: VAST Challenge 2017 Mini-Challenge
1*. IEEE Conference on Visual Analytics Science and Technology.
Award for Actionable and Detailed Analysis.

Paas, K. J. (2024). *Take Home Exercise 3: VAST Challenge 2024
Mini-Challenge 3*. ISSS608 AY2023-24, Singapore Management
University.
[https://isss608-kjcpaas.netlify.app/take-home_exs/ex3/take-home_ex3](https://isss608-kjcpaas.netlify.app/take-home_exs/ex3/take-home_ex3)

Cheong, H. S. (2023). *Take Home Exercise 3*. ISSS608 AY2022-23,
Singapore Management University.
[https://isss608-ay2023-haileycsy.netlify.app/take-home_ex/take-home_ex03/take-home_ex03](https://isss608-ay2023-haileycsy.netlify.app/take-home_ex/take-home_ex03/take-home_ex03)

Fong, B. X. (2023). *Take Home Exercise 3*. ISSS608 AY2022-23,
Singapore Management University.
[https://fongbx-isss608-visual-analytics.netlify.app/take-home%20ex/take-home_ex03/take-home_ex03](https://fongbx-isss608-visual-analytics.netlify.app/take-home%20ex/take-home_ex03/take-home_ex03)

Kwa, K. B. (2023). *Take Home Exercise 3*. ISSS608 AY2022-23,
Singapore Management University.
[https://cosmic-boon.netlify.app/take-home_ex/take-home_ex03/take-home_ex3](https://cosmic-boon.netlify.app/take-home_ex/take-home_ex03/take-home_ex3)

Ong, C. H. (2023). *Take Home Exercise 3*. ISSS608 AY2022-23,
Singapore Management University.
[https://smu-isss608-vaa.netlify.app/take-home_ex/take-home_ex03/take-home_ex03](https://smu-isss608-vaa.netlify.app/take-home_ex/take-home_ex03/take-home_ex03)

Seng, J. Y. (2024). *Take Home Exercise 3*. ISSS608 AY2023-24,
Singapore Management University.
[https://jingyi-vaa.netlify.app/take-home_ex/take-home_ex03/take-home_ex03_eda](https://jingyi-vaa.netlify.app/take-home_ex/take-home_ex03/take-home_ex03_eda)

## Session Information

For full reproducibility, the R session used to render this
report is shown below.

```{r}
sessionInfo()
```